Hammerhead-type FXR agonists induce an eRNA FincoR that ameliorates nonalcoholic steatohepatitis in mice

Version 1

Abstract

Abstract

The nuclear receptor, Farnesoid X Receptor (FXR/NR1H4), is increasingly recognized as a promising drug target for metabolic diseases, including nonalcoholic steatohepatitis (NASH). Protein coding genes regulated by FXR are well known, but whether FXR also acts through regulation of expression of long non-coding RNAs (lncRNAs), which vastly outnumber protein-coding genes, remains unknown. Utilizing RNA-seq and GRO-seq analyses in mouse liver, we found that FXR induces many RNA transcripts from chromatin regions bearing enhancer features. Among these we discovered a previously unannotated liver-enriched enhancer-derived lncRNA (eRNA), termed FincoR. We further show that FincoR is specifically induced by hammerhead-type FXR agonists, including GW4064 and tropifexor. CRISPR/Cas9-mediated liver-specific knockdown of FincoR in dietary NASH mice reduced the beneficial effects of tropifexor, an FXR agonist currently in clinical trials for NASH and primary biliary cholangitis (PBC), indicating that that amelioration of liver fibrosis and inflammation in NASH treatment by tropifexor is mediated in part by FincoR. Overall, our results indicate that pharmacological activation of FXR by hammerhead-type agonists induces a novel eRNA, FincoR, ameliorating NASH in mice. FincoR may represent a new drug target for combating metabolic disorders, including NASH.

Introduction

Nonalcoholic fatty liver disease (NAFLD) is the most common chronic liver disease and a leading cause of liver transplants and liver-related death (Friedman and Sanyal 2018). NAFLD begins with simple steatosis but may further progress to a severe form, non-alcoholic steatohepatitis (NASH), and later, fatal cirrhosis and liver cancer (Friedman and Sanyal, 2018). Despite its striking global increase and clinical importance, there is no approved drug for NASH. The urgent need for development of therapeutic agents for NASH has greatly increased research interest in the nuclear receptor, Farnesoid X Receptor (FXR, NR1H4) (Evans and Mangelsdorf, 2014).

FXR is activated by its physiological ligands, bile acids (BAs), and regulates expression of genes involved in BA, lipid, and glucose metabolism and hepatic autophagy, which maintains metabolite levels and metabolic homeostasis (Lee et al., 2006; Calkin and Tontonoz, 2012; Lee et al., 2014; Seok et al., 2014; Kliewer and Mangelsdorf, 2015). Ligand-activated FXR also protects against hepatic inflammation and liver injury (Wang et al., 2008; Jung et al., 2020). The action of FXR, similar to other nuclear receptors, is achieved primarily by binding to chromatin at specific gene loci to regulate the transcription of target genes (Lee et al., 2006; Thomas et al., 2010; Lee et al., 2012; Calkin and Tontonoz, 2012). Consistent with its crucial physiological functions, FXR is increasingly recognized as a promising drug target, particularly for liver diseases, such as NASH and primary biliary cholangitis (PBC) (Downes et al., 2003; Ali et al., 2015; Abenavoli et al., 2018; Kremoser, 2021). For example, semi-synthetic or non-steroidal synthetic agonists of FXR, including obeticholic acid (OCA) and hammerhead-type agonists, such as tropifexor and cilofexor, are currently under clinical trials for NASH and PBC patients (Tully et al., 2017; Abenavoli et al., 2018; Kremoser, 2021; Sanyal et al., 2023). However, how pharmacological activation of FXR mediates such beneficial therapeutic effects is poorly understood.

Non-protein-coding RNAs (ncRNAs) are one of the fascinating discoveries of modern biology (Cech and Steitz, 2014). Indeed, only about 2% of the human genome encodes proteins, while the vast majority are transcribed into ncRNAs (The ENCODE Project Consortium, 2012). Of these ncRNAs, long non-coding RNAs (lncRNAs) are a group of transcripts longer than 200 nucleotides and play important roles in diverse biological processes (Lam et al., 2014; Li et al., 2016; Sartorelli and Lauberth, 2020; Statello et al., 2021; Mattick et al., 2023). Many lncRNAs are produced from genomic regions bearing epigenetic features of enhancers (Hon et al., 2017; Mattick et al., 2023). This is consistent with the idea that many transcriptional enhancers actively transcribe noncoding RNAs that are referred to as eRNAs, some of which are functionally important for enhancer functions (Lai and Shiekhattar, 2014; Li et al., 2016). Due to the increasing technical sensitivity of RNA detection, many enhancers have been found to produce eRNAs (>85k in humans and >57k in mice) (Hirabayashi et al., 2019). Many eRNAs have not yet been annotated in current lncRNA databases, such as GENCODE. Exploring eRNA landscapes and functions in diverse biology and disease states will facilitate our understanding of both lncRNAs and enhancers (Li et al., 2016; Mattick et al., 2023). However, the landscape of eRNAs in mouse liver has been minimally explored (Fang et al., 2014), and has not been studied in response to specific nuclear receptor activation, such as by FXR. Moreover, functional role of eRNAs in vivo in intact organisms is understudied.

FXR was shown to regulate expression of small non-coding microRNA (miR) genes, for example miR-34a and miR-802 (Lee et al., 2010; Seok et al., 2021), but it has not been reported whether FXR mediates its function through regulating lncRNAs, which far outnumber miRs. Because FXR directly regulates expression of its target genes (Thomas et al., 2010; Lee et al., 2012), we examined if FXR regulates enhancer-derived lncRNAs, and if such transcripts participate in its physiological or pharmacological functions. By utilizing RNA-seq and GRO-seq analyses of mouse liver treated by FXR ligands, we identified a series of FXR-regulated eRNAs, including many that were previously unannotated. Among these, we focused on a highly induced and abundantly expressed eRNA that we referred to as FXR-induced non-coding RNA (FincoR) for functional studies. FincoR is highly enriched in mouse liver and is induced specifically by hammerhead-type FXR agonists, including GW4064 and tropifexor. In vivo studies utilizing CRISPR/Cas9-mediated liver-specific knockdown of FincoR in dietary NASH mice indicated that FincoR is critically involved in mediating the beneficial pharmacological effects of tropifexor in reducing liver fibrosis and inflammation.

Results

Activation of FXR by GW4064 induces a novel eRNA, FincoR, in mouse liver

To identify enhancer-derived lncRNAs (eRNAs) potentially regulated by FXR, we first obtained a list of putative enhancers in mouse liver based on ENCODE H3K27ac ChIP-Seq data (ENCFF001KMI, see methods). Then, we performed ribo-depleted total RNA-seq in the livers from mice treated with a specific FXR agonist, GW4064, to identify transcripts produced from these enhancer regions (see Supplementary Table S1). To avoid confounding issues of RNA signals from genes, we specifically focused on intergenic enhancer regions (+/-3kb from H3K27ac peak center) that harbor discernible RNA-Seq signals (RPKM > 1).

This genomic analysis resulted in identification of 190 high-confidence eRNAs in mouse liver. Among these, 14 eRNAs were upregulated and 5 were downregulated by GW4064 treatment (FDR < 0.05, Log2 FC > 1, Figure 1A) (see Supplementary Table S2). FXR-regulated eRNAs were produced adjacent to many genes with important roles in liver metabolism and disease, for example, Hes1 (Figure 1-Figure Supplement 1A, B). One of the most robustly induced eRNAs was FXR-induced non-coding RNA (FincoR), an unannotated novel transcript located on chromosome 19 (Figure 1A, B, C). Reverse transcription qPCR (RT-qPCR) confirmed that treatment with GW4064 substantially induced expression of FincoR, more than 10-fold in mouse liver, which is similar to induction of Shp, a well-known FXR target gene (Figure 1D) (Goodwin et al., 2000; Claudel et al., 2005; Evans and Mangelsdorf, 2014). Induction of FincoR by GW4064 was transient, peaked within 1 h and then, declined gradually, similar pattern to Shp (Figure 1-Figure Supplement 1C). Expression of genes adjacent to FincoR, including Gcnt1, Rfk, Pcsk5 and Prune2, did not change after acute 1 h FXR activation as shown by RNA-seq (Figure 1C), and confirmed for Gcnt1 by time course qPCR (Figure 1-Figure Supplement 1C).

Activation of FXR by GW4064 induces FincoR, a novel eRNA, in mouse liver.

(A) C57BL/6 male mice were fasted overnight and i.p. injected with vehicle or GW4064 (30 mg/kg) for 1 h, and livers were collected (n = 4 mice). RNAs from two mice (per group) were pooled for RNA-seq. Volcano plot showing significantly induced eRNAs (FincoR highlighed) by GW4064 treatment. The x axis denotes log2 fold change (GW4064/Veh) of eRNAs and the y axis denotes -log10 FDR of eRNAs. (B) The bar plot showing the RPMs of up-regulated eRNAs by GW4064 treatment. (C) FincoR induction by GW4064 treatment was shown in the genome browser track. (D) GW4064 induction of FincoR and Shp was measured in the liver by qPCR (n = 4 mice). Data are presented as mean ± SEM. Statistical significance was determined by the two-way ANOVA Sidak’s multiple comparisons test with *p < 0.05 and ***p < 0.001. Veh, vehicle; GW, GW4064; pos, positive strand; neg, negative strand; Refseq, reference sequence.

We also found that the short GW4064 treatment resulted in 590 up-regulated genes and 500 down-regulated genes (FDR < 0.05) (Figure 1-Figure Supplement 2A). GO enrichment analysis of these differentially expressed genes (DEGs) revealed their roles in the regulation of triglyceride, fatty acid, and cholesterol metabolism (Figure 1-Figure Supplement 2B), which is consistent with known roles of FXR in these physiological processes (Claudel et al., 2005).

Ligand-activated FXR directly activates transcription of eRNAs, including FincoR

We sought to examine, 1) if these eRNAs were directly activated by FXR, and 2) if the activation takes place transcriptionally using FXR liver-specific knockout (FXR-LKO) mice that were treated with GW4064 to determine if FXR was required for induction of eRNAs by GW4064. FXR-LKO mice were generated from FXR floxed mice (FXR-flox) (Figure 2-Figure Supplement 1A) and transcription of eRNAs was detected by global run-on sequencing (GRO-seq), a widely used method to detect nascent RNA transcription, including eRNAs (Lam et al., 2014; Li et al., 2016; Sartorelli and Lauberth, 2020). GRO-seq showed that FXR-induced eRNAs were activated transcriptionally by GW4064 in FXR-flox mice, but such induction was abolished in FXR-LKO mice (Figure 2A). In particular, FincoR is robustly induced in the GRO-Seq analysis and its induction is dependent on hepatic FXR (Figure 2B). RT-qPCR confirmed the FXR-dependent expression of FincoR, similar to that of Shp (Figure 2-Figure Supplement 1B).

Ligand-activated FXR directly activates transcription of eRNAs including FincoR in the liver.

(A) Experimental outline: FXR-Flox and LKO male mice were fasted overnight and treated with vehicle or GW4064, livers were collected 1 h later (n = 2 each) and nuclei was isolated for Gro-seq. Box plot shows the up-regulated eRNA GRO-Seq RPKMs in different conditions. (B) Gro-seq profile of FincoR expression and genome browser tracks of H3K4me1, H3K27ac, FXR and RXRα binding peaks at FincoR locus were shown. (C) ChIP assays were performed in liver samples from Fig. 1A to detect FXR, RXRα, and BRD4 occupancy to the FXR binding peak region close to the transcription start site of FincoR. (D) HepG2 cells were transfected with luciferase reporter expressing wild type FXRE or mutant FXRE (see materials and methods) for 24 hrs before further subject to GW4064 for 6 hrs. The relative luciferase activity was shown.

We next examined if FXR binds to these enhancers that produce the identified eRNAs by analyzing published mouse liver ChIP-Seq data for FXR (Supplementary Table S3) (Thomas et al., 2010; Lee et al., 2012). FXR binding was strongly enriched at the enhancers associated with FXR-induced eRNAs as were the enhancer marks H3K27ac and H3K4me1 (Figure 2-Figure Supplement 1C). In Figure 2B, the results show the binding of FXR and the presence of histone marks at the FincoR enhancer region determined by ChIP-seq as compared with nascent transcripts detected by Gro-seq. These analyses support the conclusion that activation of FXR transcriptionally induces this series of eRNAs via chromatin binding at these enhancers, including FincoR. We validated FXR binding at the enhancer region that produces FincoR using mouse liver ChIP (Figure 2C). We also examined binding of a well-known DNA binding partner of FXR, Retinoid x Receptor alpha (RXRα/NR2B1) (Evans and Mangelsdorf, 2014; Zheng et al., 2018), and bromodomain-containing protein 4 (BRD4), an acetylated histone reader protein that often binds at active enhancers (Chen et al., 2016; Li et al., 2016; Rahnamoun et al., 2018; Sartorelli and Lauberth, 2020) and a transcriptional coactivator of FXR (Jung et al., 2020). GW4064 treatment resulted in substantial increases in recruitment of both FXR and RXRα to the enhancer region close to the transcription start site of FincoR (arrow shown below in Figure 2B), whereas binding was not detected at a control region (Figure 2C). We also found that BRD4 occupancy was increased at this enhancer region after GW4064 treatment (Figure 2C). These results indicate that GW4064 activation of FXR leads to increased occupancy of the FXR/RXRα heterodimer and BRD4 to the enhancer region to upregulate FincoR eRNA in the liver.

We identified an inverted repeat1 (IR1) motif that was known to bind FXR (Lee et al., 2006; Calkin and Tontonoz, 2012) within the major FXR binding peak near the start site of FincoR (arrow shown below in Figure 2B), which we also refer to as FXRE (Figure 2B, C). We examined the functionality of this IR1 motif for mediating transcriptional activation by GW4064 using reporter assays (Figure 2D). We cloned the region containing the IR1 motif into the pGL4.23 luciferase reporter and generated a mutated IR1 motif construct as a comparison (Figure 2D). After transfection into human hepatic HepG2 cells, GW4064 treatment significantly elevated the luciferase activity of the reporter with the wild type IR1 motif, but not with the mutated IR1 motif (Figure 2D). Together, these results suggest that GW4064-activated FXR directly upregulates FincoR expression.

FincoR is a liver-specific nucleus-enriched eRNA

Because enhancers and eRNAs generally act in a tissue-specific manner (Li et al., 2016; Sartorelli and Lauberth, 2020), we examined the tissue-specific expression of FincoR in mice. Strikingly, FincoR is highly expressed in the liver and it is expressed at extremely low levels in most other tissues, except for a detectable, but still fairly low, level in the lung. GW4064 treatment resulted in induction of FincoR specifically in the liver (Figure 3A). The level of FincoR detected in isolated primary mouse hepatocytes (PMHs) was similar to that in mouse liver tissue treated with GW4064, suggesting that the majority of FincoR is present in hepatocytes (Figure 3B).

FincoR is a liver-specific nucleus-enriched eRNA.

(A) C57BL/6 male mice were fasted overnight and i.p. injected with vehicle or GW4064 (30 mg/kg) for 1 h. Tissues were collected and FincoR mRNA level was measured (n = 2 mice). (B) C57BL/6 male mice were fasted overnight and i.p. injected with vehicle or GW4064 (30 mg/kg) for 1 h. One small piece of liver was snap-frozen for later RNA isolation and the remaining part was used for immediate primary hepatocyte isolation. Then, RNAs were extracted from liver or primary hepatocytes and FincoR expression was measured. (C) Agarose gel electrophoresis of PCR products generated in 5′ (left) and 3′ (right) RACE of FincoR in liver samples. Primer locations were show. (D) Schematic diagrams of full-length FincoR. (E) PhyloCSF analysis of the coding potential of FincoR. (F) In vitro translation of FincoR using the Promega Transcend Non-Radioactive Translation Detection Systems. Luciferase is used as controls for coding RNA. (G) qPCR analysis of FincoR, 36b4 in Poly(A)+ and Poly(A)-RNA fractions from GW4064 treated mouse liver. (H) FincoR identified in the subcellular fraction using cellular fractionation assays. The primary hepatocytes were isolated from GW4064 or DMSO treated mouse liver and the cytoplasm and nucleus fractions of these hepatocytes were separated and both fractions were subjected to RNA extraction and qPCR. RACE, rapid amplification of cDNA ends.

By using the 5’ and 3’ rapid amplification of cDNA ends (RACE), we identified one transcript of FincoR that is of approximately 3.7 kb in length (Figure 3C). However, based on RNA-seq, the length of FincoR is over 10 kb (Figure 1C), suggesting there are likely additional multiple RNA isoforms that we were not able to identify by RACE. We next analyzed the coding potential of FincoR utilizing a comparative genomic program, PhyloCSF. While adjacent genes Gcnt1 and Prune2 (Figure 3D, E) were correctly predicted to encode proteins, FincoR did not contain a potential protein-coding open reading frame (Figure 3E). Consistent with this bioinformatic prediction, a vector expressing FincoR failed to produce any proteins in an in vitro transcription/translation assay (Figure 3F) confirming that the FincoR transcript is a noncoding RNA. FincoR transcripts were enriched by binding to oligo-dT beads, suggesting that FincoR is 3’ polyadenylated (Figure 3G). Further, FincoR was detected in the nuclear compartment and GW4064 treatment increased the nuclear abundance of FincoR (Figure 3H), consistent with its potential transcriptional regulatory function. Together, these results from molecular biochemical characterization studies reveal that FincoR is a liver-enriched nuclear polyadenylated eRNA.

FincoR is induced specifically by the hammerhead-type synthetic FXR agonist

To determine whether induction of the FXR-induced FincoR is ligand-specific, we examined the effects of hammerhead-type synthetic FXR agonists, GW4064, cilofexor, and tropifexor; a semi-synthetic agonist, OCA; and a non-hammerhead-type gut-specific agonist, fexaramine (Figure 4A) (Downes et al., 2003; Fang et al., 2015). Remarkably, treatment with each of the hammerhead type agonists for 1 h resulted in a robust induction of FincoR (Figure 4B), whereas FincoR levels were unchanged after treatment with OCA or fexaramine for 1 h (Figure 4B). Treatment with OCA for 4 h or even one-week treatment with OCA failed to induce hepatic FincoR in mice, while expression of Shp was significantly induced (Figure 4C, D). Acute feeding with a diet supplemented with 0.5% cholic acid (CA), a primary BA, for 6 h also failed to induce FincoR expression (Figure 4E). Collectively, these results demonstrate that FincoR is induced specifically by hammerhead type FXR agonists.

FincoR is induced by the hammerhead class of non-steroidal FXR agonists, including GW4064 and tropifexor.

(A) The chemical structures of the FXR agonists including the hammerhead class of non-synthetic FXR agonists, non-hammerhead-type non-synthetic agonists, semi-synthetic BA and natural BA. (B) C57BL/6 mice were treated with GW4064 (30 mg/kg) or Cilofexor (30 mg/kg) or Tropifexor (0.5 mg/kg) or Fexaramine (100 mg/kg) or OCA (20 mg/kg) for 1 h after over-night fasting. The liver RNAs were extracted and FincoR expression was measured (n = 3∼4 mice). (C) C57BL/6 mice were treated with OCA (20 mg/kg) or Fexaramine (100 mg/kg) for 4 hrs after over-night fasting. The liver RNAs were extracted and FincoR expression was measured (n = 3 mice). (D) C57BL/6 mice were daily treated with OCA (20 mg/kg) for 7 days after over-night fasting. The liver RNAs were extracted and FincoR expression was measured (n = 5 mice). Shp gene was used as a positive control. (E) C57BL/6 mice were fed with 0.5% cholic acid (CA) diet for 6 hrs after over-night fasting. The liver RNAs were extracted and FincoR expression was measured (n = 3∼4 mice). (B-E) Data are presented as mean ± SEM. Statistical significance was determined by the Student’s t test with *p < 0.05, **p < 0.01 and ***p < 0.001.

Generation of CRISPR/Cas9-mediated FincoR liver-specific knockdown mice

To explore the functional role of hepatic FincoR, we utilized the CRISPR/Cas9 technique to generate FincoR liver-specific knockdown (FincoR-LKD) mice (Figure 5-Figure Supplement 1A). Adenoviral-mediated sgRNA expression in Cas9 mice resulted in downregulation of FincoR specifically in the liver by about 60%, but not in other tissues (Figure 5B). As FXR is a key regulator of bile acid, cholesterol, lipid, and glucose metabolism and FincoR is specifically regulated by FXR, FincoR may have a role in the metabolic process in physiology and disease. We examined liver triglyceride, cholesterol, bile acid, glycogen and serum non-esterified fatty acids (NEFA) but FincoR downregulation did not result in any significant changes under physiological conditions (Figure 5C).

Generation of CRISPR/Cas9-mediated FincoR liver-specific knockdown mice.

(A) Male Cas9 mice were infected with adenovirus expressing sgRNA for FincoR or control for 1 week. Then the liver and serum were collected from these mice after 4∼5 hrs of fasting. (B) The expression of FincoR in the liver was measured by qPCR (n = 4 mice). (C) Hepatic triglyceride, cholesterol, bile acid, glycogen and serum NEFA were measured (n = 5∼7 mice) (D) Male Cas9 mice were infected with adenovirus expressing sgRNA for FincoR or control for 1 week. Then these mice were fasted overnight and treated with GW4064 for 3 hrs before tissue collection. RNA-seq profiles of expression of hepatic FincoR and the adjacent genes were shown (n = 2 mice). (E) Genome-wide changes in mRNA expression shown in a volcano plot. The numbers refer to the number of genes up-or down-regulated by 2-fold or more with a p-value < 0.01. (F) Gene ontology analysis of biological pathways using DAVID Tools.

To explore the molecular signatures and pathways affected by FincoR, we examined the global gene expression by RNA-seq analysis in mouse liver after FincoR knock-down (Figure 5D). While FincoR was markedly downregulated, the neighboring genes, such as Gcnt1, Rfk, Pcsk5, were largely unchanged (Figure 5D). The RNA-seq analysis revealed 18 up-regulated genes and 53 downregulated genes in FincoR-LKD liver (Supplementary Table S4; Figure 5E). GO analysis indicated that these genes were enriched in pathways involved in fatty acid oxidation, organelle organization and metabolic process (Figure 5F). Among the down-regulated genes, Ppp1r3g which has a role in controlling glycogen synthesis, and Igfbp2 which functions in insulin resistance, were markedly reduced (Supplementary Table S4; Figure 5-Figure Supplement 1D, E). Among the up-regulated genes, expression of Eda2r was substantially increased, which belongs to a member of tumor necrosis factor receptor superfamily and is involved in inflammation, the immune response and development (Figure 5-Figure Supplement 1F). Fndc1 which is involved in fibronectin matrix remodeling was also suppressed by FincoR (and thus upregulated upon its KO, Figure 5-Figure Supplement 1G). These studies suggested that FincoR may have a role in modulating metabolic homeostasis by regulating genes involved in metabolism and inflammation.

Amelioration of hepatic steatosis mediated by tropifexor is independent of FincoR in diet-induced NASH mice

Tropifexor, also known as LJN452, is a highly potent hammerhead type FXR agonist that is currently under clinical trials for NASH and PBC patients (Tully et al., 2017; Kremoser, 2021; Sanyal et al., 2023). Because FincoR is induced specifically by the hammerhead class of FXR agonist (Figure 4) and has a potential role in the regulation of metabolism and inflammation (Figure 5), we hypothesized that FincoR may play a role in tropifexor-mediated beneficial effects on reducing NASH pathologies in mice.

We adopted a mouse model that had been fed the Amylin Liver NASH-promoting (AMLN) diet (Zhao et al., 2018; Hernandez et al., 2019; Sun et al., 2022), and examined the effects of liver-specific downregulation of FincoR in potentially impacting tropifexor effects on NASH pathology. Cas9 mice fed AMLN diet for 12 weeks were injected via tail veins with adenovirus expressing control sgRNA or FincoR sgRNA, respectively, and then treated daily with tropifexor (0.3 mg/kg) for 12 days (Figure 6A). In these mice, as a technical validation of RNA induction and knockdown, FincoR levels were significantly increased by FXR agonist tropifexor, and the increase was blocked by adenovirus expressing sgRNA for FincoR (Figure 6B).

In diet-induced NASH mice, tropifexor-mediated beneficial effects on reducing hepatic steatosis are largely independent on FincoR.

(A-E) Male Cas9 mice were fed with NASH diet for 12 weeks. Then these mice were randomly assigned to 3 groups and infected with adenovirus expressing sgRNA for FincoR or control. Three days later, the mice were treated with Tropifexor (0.3 mg/kg) for 12 days. On the tissue collection day, the mice were given the last shot of Tropifexor or vehicle and fasted for 4 hrs before tissues were collected (n = 6 ∼ 7 mice). (A) Experimental scheme. (B) Hepatic FincoR expression was measured (n = 6∼7 mice). (C) Oil Red O staining of liver sections. Scale bar (50 μm). (D) Hepatic TG, hepatic cholesterol, gallbladder BA and hepatic BA levels were measured (n = 6 ∼ 7 mice). (E) mRNA levels in the liver of the indicated genes involved in bile acid regulation and lipid regulation (n = 6 ∼ 7 mice). (B, D and E) Data are presented as mean ± SEM. Statistical significance was determined by the one-way ANOVA (Sidak’s multiple comparisons test) with *p < 0.05, **p < 0.01 and ***p < 0.001. Ad, adenovirus; H & E, hematoxylin and eosin; TG, triglyceride; Veh, vehicle; ns, not significant.

We then examined the effect of tropifexor treatment and FincoR downregulation on hepatic steatosis in these mice. Tropifexor treatment markedly reduced neutral lipids determined by Oil Red O staining of liver sections (Figure 6C) and liver TG levels (Figure 6D), and these beneficial effects on reducing fatty liver were not altered by FincoR downregulation. Also, FincoR downregulation had little effects on liver cholesterol and gallbladder bile acid levels, although these levels were reduced by Tropifexor (Figure 6D).

Consistent with the phenotypes, hepatic expression of key genes involved in BA synthesis was dramatically reduced by tropifexor treatment (i.e., Cyp7a1 and Cyp8b1), which is consistent with decreased gallbladder BA levels by this drug (Figure 6E). Similarly, tropifexor also lowered lipid synthesis genes (Srebp1, Lipin, Scd1), consistent with decreased liver TG levels. However, downregulation of FincoR did not result in changes in mRNA levels of these genes (Figure 6E). These results indicate that tropifexor-mediated beneficial effects on reducing hepatic steatosis are independent of FincoR.

FincoR facilitates alleviation of liver inflammation by tropifexor in diet-induced NASH

Tropifexor ameliorated fibrotic NASH pathologies in preclinical studies (Tully et al., 2017; Hernandez et al., 2019) and recently completed phase 2 clinical trials for NASH patients (Sanyal et al., 2023). We, therefore, further examined the effects of FincoR downregulation on altering other NASH pathologies, including hepatocellular apoptosis, liver fibrosis and inflammation.

In the same AMLN diet-fed mice above (Figure 6), analyses of liver sections revealed that tropifexor treatment reduced hepatocyte swelling/ballooning (H&E staining), decreased numbers of apoptotic cells (TUNEL staining), alleviated fibrosis (Sirius red staining), and lowered infiltration of macrophages (F4/80 staining) (Figure 7A). Remarkably, these tropifexor-mediated beneficial effects on NASH pathologies were all markedly diminished by FincoR downregulation (Figure 7A). This was consistently found in various liver lobes and two representative pictures from 2 different lobes are shown (Figure 7A). In control experiments, FincoR downregulation in vehicle-treated mice did not result in marked changes in NASH pathologies (Figure 7-Figure Supplement 1). These results demonstrate that FincoR is required for tropifexor-mediated beneficial effects on reducing NASH pathologies, specifically reducing hepatic inflammation, fibrosis and hepatocyte apoptosis. Consistent with these results, serum ALT levels, an indicator of liver damage, were significantly elevated after FincoR downregulation (Figure 7B). Protein levels of key inflammatory markers, IL1β and CCL2, in liver extracts were also elevated after FincoR downregulation (Figure 7C).

In diet-induced NASH mice, tropifexor-mediated beneficial effects on reducing liver fibrosis and inflammation are diminished by FincoR downregulation.

H & E, F4/80, Sirius Red and TUNEL staining of liver sections from Figure 6. Scale bar (50 μm). Image analyses were done using Image J and the area of collagen staining, TUNEL and F4/80 levels were quantified (n = 5/group). (B) Serum ALT level was measured (n = 5/group). (C) IL-1β and CCL2 levels in the liver tissues were determined by ELISA (n = 5/group). (D) mRNA levels in the liver of the indicated genes involved in inflammation, fibrosis and cell death (n = 5/group). (E) Model: FincoR is a liver-specific lncRNA that is induced specifically by hammerhead-type FXR agonists (top). In diet-induced NASH mice, FincoR is required for tropifexor-mediated beneficial effects on reducing hepatic inflammation, fibrosis, and cell death with the mechanisms to be explored further (bottom). (A-D) Data are presented as mean ± SEM. Statistical significance was determined by the one-way ANOVA (Sidak’s multiple comparisons test) with *p < 0.05, **p < 0.01 and ***p < 0.001.

Consistent with the phenotypes from histological analyses, hepatic expression of fibrosis and inflammation was altered by FincoR knockdown. For example, tropifexor treatment reduced mRNA levels of several genes that promote fibrosis (Col1a1, Col1a2, Acta2) and hepatic inflammation (Eda2r, Ifng, Ccl3), whereas these reductions were largely reversed by FincoR downregulation (Figure 7D). We also detected increased expression of inflammatory genes (Ccl2, Ccr2, Lcn2) and extracellular matrix remodeling gene (Fndc1) in mice liver with FincoR knockdown (Figure 7D). Tropifexor treatment suppresses hepatic apoptosis by reducing pro-apoptotic genes (Ctsb, Ctss) and upregulating anti-apoptotic genes such as Bcl2 (Warren et al., 2019). Importantly, these effects were significantly reversed by FincoR downregulation (Figure 7D).

Collectively, these results demonstrate that in diet-induced NASH mice, pharmacological activation of FXR by tropifexor reduced fibrosis, apoptosis, and inflammation in NASH mice, which was dependent, at least in part, on the induction of FincoR.

Discussion

FXR maintains metabolic homeostasis by transcriptional regulation of genes. Direct regulation of protein-coding genes by FXR, including Shp, is well characterized. In this study we show that FXR also mediates its functions by induction of lncRNA genes, which vastly outnumber protein-coding genes. We further show that pharmacological activation of FXR by hammerhead-type agonists induces a liver-specific enhancer-derived lncRNA, which we named FincoR, that contributes to reduction of NASH pathologies in mice.

FincoR is specifically induced by the hammerhead class of FXR agonists, such as GW4064, cilofexor, and tropifexor. It was not induced by other FXR ligands, including its endogenous ligand, cholic acid, suggesting that FincoR does not contribute to physiological functions of FXR, but instead plays a role in pharmacological FXR responses and possibly in pathological processes. GW4064 is the mother compound of these isoxazole type ligands but is not an ideal therapeutic agent because of its poor water solubility and pharmacokinetics (Abel et al., 2010). In contrast, tropifexor and cilofexor have better pharmacokinetics and were generally well-tolerated in clinical trials for NASH and PBC patients (Tully et al., 2017; Abenavoli et al., 2018; Kremoser, 2021; Sanyal et al., 2023), but the underlying mechanisms for their beneficial effects are poorly understood. Intriguingly, a recent study showed that the gene signature regulated by tropifexor-activated FXR appears to be broader than that of OCA, partly because the tropifexor backbone allows a more favorable interaction of FXR with coactivators or epigenomic modulators (Hernandez et al., 2019). Further, tropifexor was shown to regulate distinct sets of genes, particularly those promoting fibrosis, inflammation, and oxidative stress, in experimental NASH (Hernandez et al., 2019). Utilizing CRISPR/Cas9-mediated liver-specific knockdown of FincoR in diet-induced NASH mice, we demonstrate that beneficial effects on reducing liver fibrosis, inflammation, and apoptosis mediated by tropifexor were largely dependent on FincoR (Model, Figure 7E).

The roles of regulatory RNAs in liver function and diseases and as therapeutic targets or drugs are increasingly being appreciated (Sallam et al., 2016; Zhao et al., 2018; Brocker et al., 2020; Li et al., 2021). For example, a critical role for an oxysterol nuclear receptor LXR-induced lncRNA, LeXis, in feedback modulation of cholesterol biosynthesis has been shown (Sallam et al., 2016). Recently, the role of a lncRNA, Pair, in the liver phenylalanine metabolism has been demosntrated (Li et al., 2021). Enhancer RNAs are a less characterized class of lncRNAs and are highly associated with enhancer functions in gene regulation (Li et al., 2016). Numerous studies have revealed transcriptional roles for eRNAs in various cellular processes (Lai and Shiekhattar, 2014; Lam et al., 2014; Li et al., 2016) but most previous eRNA studies have used culture cells (Li et al., 2013; Hsieh et al., 2014), with only a few in vivo studies in mouse models (Mirtschink et al., 2019; Tang et al., 2023). In our current study, through integrative analysis of transcriptome and histone mark ChIP-Seq, we identified FXR-induced eRNAs, including the highly induced FincoR. We further examined the expression of function of FincoR utilizing mouse models. Our current work characterized the role of FincoR in gene regulation and in mediating beneficial pharmacological effects of FXR agonists in NASH, representing an important progress in understanding physiological/pathological roles of eRNAs in vivo. Future work is warranted to understand the exact mechanisms by which FincoR facilitates action of FXR agonists to alleviate inflammation, fibrosis and apoptosis.

While this study focused on regulation of FincoR by pharmacological activation of FXR, physiological and pathological regulations of FincoR appear to be complex. FincoR was not induced by acute cholic acid feeding, but FincoR levels were elevated in mouse models of cholestatic liver (data not shown), where BA metabolism is dysregulated. Since different BAs can activate or repress the gene-regulating function of FXR (Wahlstrom et al., 2016), altered BA composition in cholestatic liver may contribute to induction of FincoR. Further, binding peaks for multiple nuclear receptors, FXR, LXR, PPARα, RXRα, and HNF-4α, were detected in the FincoR gene locus so that regulation of FincoR likely involves combinatorial regulation by multiple nuclear receptors (Other Supplementary Figure 1). Interestingly, occupancy of the nuclear receptor PPARα at the enhancer region was increased in fasted mice (Other Supplementary Figure 2). It will be interesting to see whether expression of FincoR is differently regulated by these nuclear receptors in response to physiological and pathological cues.

FXR is increasingly recognized as an important therapeutic target for enterohepatic diseases, but the development of clinically applicable and more targeted FXR-based therapy is still challenging. In this study, we provided the first characterization of an eRNA, FincoR, induced by pharmacological activation of FXR and show that FincoR has a beneficial role in reducing liver fibrosis and inflammation in dietary NASH mice. Complete understanding of the function and mechanisms of FincoR may provide novel insights for the development of desirable therapy for NASH and other chronic liver diseases.

Materials and Methods

Animal experiments

All animal studies were performed according to procedures approved by the Institutional Animal Care & Use Committee at the University of Illinois at Urbana-Champaign and were in accordance with National Institutes of Health guidelines. Mice were maintained in 12/12 h light/dark cycles and fed standard rodent chow. FXR-LKO mice were generated by breeding FXR floxed mice with Albumin-Cre mice (The Jackson Lab). FincoR-LKD mice were generated as previously reported (Zhao et al., 2018). Briefly, Cas9 transgenic mice (JAX #024858) were via the tail vein injected with adenoviruses (approximately 5x108 PFU) expressing two sgRNAs targeting FincoR (sgRNA1: GGGTTAAGAGCTGTAGGCTG and sgRNA2: ACTTCTATGTCCAACAACCG). The sequences of sgRNAs were designed using a CRISPR design tool (http://crispr.mit.edu/).

Mice were given a single dose of vehicle or 30 mg/kg GW4064 (in corn oil, Tocris Bioscience, #2473) after overnight fasting. Mice were treated with 0.5 mg/kg tropifexor (in corn oil, MedChem Express, HY-107418), 30 mg/kg cilofexor (in corn oil, MedChemExpress, HY-109083), 100 mg/kg fexaramine (in 0.5% methylcellulose, MedChem Express, HY-10912), and 20 mg/kg OCA (in 0.5% methylcellulose, MedChem Express, HY-12222) as indicated. C57BL6 mice were fed with a chow diet containing 0.5% CA for 6 hrs (Jung et al., 2020).

For liver-specific downregulation of FincoR, mice were fed AMLN diet (Research Diets, D09100310, 40 kcal% fat, 2% cholesterol, 20 kcal% fructose) for 12 weeks and then, were injected with adenovirus expressing control sgRNA or sgRNA targeting FincoR and administration of tropifexor was started 3 days later given at 0.3 mg/kg dissolved in corn oil.

RNA-Seq

C57BL/6 mice were fasted overnight and treated with 30 mg/kg GW4064 or vehicle for 1 h. RNA-seq was performed as previously described (Byun et al., 2018; Seok et al., 2018; Byun et al., 2020). Briefly, the RNAs from livers (n = 2 mice/group, each group contains RNAs pooled from 2 mice) were prepared (RNeasy, Qiagen). Ribosomal RNA was removed with the Ribozero HMR Gold kit (Illumina).

Construction of strand-specific RNAseq libraries

Construction of the RNAseq libraries and sequencing on the Illumina NovaSeq 6000 were performed at the Roy J. Carver Biotechnology Center at the University of Illinois at Urbana-Champaign. After DNase digestion, purified total RNAs were analyzed on a Fragment Analyzer (Agilent) to evaluate RNA integrity. The total RNAs were converted into individually barcoded polyadenylated mRNAseq libraries with the Kapa HyperPrep mRNA kit (Roche). Libraries were barcoded with Unique Dual Indexes (UDI’s) which have been developed to prevent index switching. The adaptor-ligated double-stranded cDNAs were amplified by PCR for 8 cycles with the Kapa HiFi polymerase (Roche). The final libraries were quantitated with Qubit (ThermoFisher) and the average cDNA fragment sizes were determined on a Fragment Analyzer. The libraries were diluted to 10 nM and further quantitated by qPCR on a CFX Connect Real-Time qPCR system (Bio-Rad) for accurate pooling of barcoded libraries and maximization of number of clusters in the flowcell.

Sequencing of libraries in the NovaSeq

The barcoded RNAseq libraries were loaded on one SP lane on a NovaSeq 6000 for cluster formation and sequencing. The libraries were sequenced from one end of the fragments for a total of 100 bp. The fastq read files were generated and demultiplexed with the bcl2fastq v2.20 Conversion Software (Illumina, San Diego, CA). The quality of the demultiplexed fastq files was evaluated with the FastQC software, which generates reports with the quality scores, base composition, k-mer, GC and N contents, sequence duplication levels and overrepresented sequences.

GRO-seq

To harvest the nuclei from mouse liver cells, the liver was harvested at indicated time and washed with a cold swelling buffer (10 mM Tris pH 7.5, 2 mM MgCl2, 3 mM CaCl2, 2 U/ml Superase-In). The nuclei were prepared by Dounce homogenization in cold swelling buffer and filtered using the cell strainer (100 μm, BD Biosciences). Nuclei were collected by centrifugation at 400 x g for 10 min, then resuspended in the lysis buffer (swelling buffer with 10% glycerol and 1% IGEPAL) and incubated on ice for 5 min. Nuclei were washed twice with the lysis buffer and resuspended at a concentration of 108 nuclei/ml in the freezing buffer (50 mM Tris pH 8.3, 40% glycerol, 5 mM MgCl2, 0.1 mM EDTA). We then followed our previous method (Li et al., 2013; Oh et al., 2021) to conduct nuclear run-on and GRO-Seq library preparation. Briefly, the nuclei in freezing buffer were subjected to the nuclear run-on reaction by mixing with an equal volume of run-on buffer (10 mM Tris-Cl pH 8.0, 5 mM MgCl2, 1 mM dithiothreitol (DTT), 300 mM KCl, 20 units of Superase-In, 1% sarkosyl, 500 µM ATP, GTP, Br-UTP and 2 µM CTP) for 5 min at 30 °C. The BrU-labeled run-on RNAs were extracted by TRIzol and purified by anti-BrdU agarose beads (Santa Cruz Biotech, sc-32323 AC). The run-on RNAs were then subjected to end repair by T4 PNK, poly-adenylation, and then to cDNA first strand synthesis by a custom primer (oNTI223) that allows circularization of the cDNA. This cDNA then was re-linearized by Ape1 (NEB), size selected by TBE gel and the products of the desired size were excised (∼320–350 bp) for final library prep and sequencing. GRO-Seq samples were run on a NextSeq 500 sequencer from Illumina with a single-end 80 nt model.

Histological analyses

For histology, tissues were dissected and immediately fixed in 10% formalin overnight and processed for paraffin embedding and H&E staining. Paraffin-embedded liver sections were incubated with F4/80 antibody, and antibody was detected using a peroxidase-based method (Abcam, ab64238). Liver collagen was detected by Sirius Red staining (Abcam, ab246832) and apoptosis was detected by TUNEL staining (Millipore, S7100). For Oil Red O staining, liver tissue was frozen in OCT compound (Sakura Finetek, 4583), sectioned, and stained. Liver sections were imaged with a NanoZoomer Scanner (Hamamatsu) and quantification was done using NIH ImageJ.

Metabolic analyses

Hepatic levels of TG (Sigma, MAK266), cholesterol (Sigma, MAK043), glycogen (Biovision, K646-100), total BA levels (Diazyme, DZ042A), serum NEFA (Sigma, MAK044) and serum ALT (Sigma, MAK052) were determined according to the manufacturer’s instructions. Mouse liver IL-1β (R&D systems, MLB00C) and CCL2 (R&D systems, DY479-05) were detected by commercially available ELISA kit.

Mouse liver ChIP

Liver ChIP assay was performed as described previously (Seok et al., 2018; Byun et al., 2020; Jung et al., 2020). Briefly, chromatin extracts were prepared from FXR floxed and FXR-LKO mouse livers after treatment of the mice with GW4064 which was followed by preclearing and immunoprecipitation using control IgG or FXR antibody (Novus Biologicals, NBP2-16550; Santa Cruz, sc-25309), RXRα antibody (Proteintech, catalog no. 21218-1-AP), or BRD4 antibody (Bethyl Laboratories, catalog #A301-985A50). Enrichment in chromatin precipitates of gene sequences was measured by qRCR using primers listed (Supplementary Table S5).

Primary mouse hepatocytes and HepG2 cells

Primary hepatocytes were isolated from C57BL/6 mice by collagenase (Worthington Biochemical Corporation, LS004188) perfusion and maintained in William’s E Medium with primary hepatocyte maintenance supplements (Gibco #CM4000) in six-well plates as described previously (Kim et al., 2015; Jung et al., 2020). HepG2 cells were maintained in DMEM supplemented with 10% fetal bovine serum at 37°C and 5% CO2.

Luciferase reporter assay

The enhancer region containing the FXR binding element was amplified by PCR and cloned into the pGL4.23 vector (Promega). HepG2 cells were transiently transfected with FXR (100 ng/well) and RXRα (5 ng/well) in combination with reporters containing the wild type FXRE or mutated FXRE (200 ng/well) (Supplementary Table S5). β-galactosidase plasmid (200 ng/well) was also transfected as internal controls. Cells were treated with GW4064 or DMSO after transient transfection. Six hours later, the cells were harvested. All reporter assays were repeated at least three times in triplicates.

RACE

5’ and 3’ RACE assays were performed using a SMARTer RACE kit (Clontech) according to the manufacturer’s instructions. The resulting PCR products were separated by electrophoresis in agarose gels and cloned into the pRACE vector provided by the kit. The transcription start sites and end sites of FincoR were determined by sequencing. The gene-specific primers used for 5’-and 3’-RACE are provided (Supplementary Table S5).

In vitro transcription and translation (TNT)

Expression plasmids for luciferase and FincoR were mixed with a Coupled Reticulocyte Lysate System (Promega). After incubating at 30°C for 60 min, translated products were separated on 4-20% gradient SDS polyacrylamide gels and transferred to PVDF membranes. Chemiluminescent detection of in vitro translated protein was performed following the manufacturer’s protocol (Promega).

RT-qPCR

Total RNA was extracted using RNeasy Mini Kit (Qiagen, 74104) and 2 µg of RNA was reverse transcribed and relative mRNA expression was normalized to those of 36B4. The qPCR primers are shown in Supplementary Tables S5.

Subcellular fractionation

Using a Cytoplasmic and Nuclear RNA Purification Kit (Norgen, Thorold, ON, Canada), the cytoplasm and nucleus fractions from primary hepatocytes isolated from livers of mice treated with GW4064 or DMSO were separated, and both fractions were subjected to RNA extraction and qRT-PCR.

Immunoblotting analysis

Total liver lysates were prepared as described before (Sun et al., 2022). Antibodies for immunoblotting for β-ACTIN (#4970) were purchased from Cell Signaling Technology. Antibodies for immunoblotting for FXR (sc-25309) were purchased from Santa Cruz.

FincoR polyadenylation study

Total RNA was prepared using the RNeasy mini prep kit (Qiagen). Poly(A)+ and poly(A)-RNA was separated by Dynabeads™ mRNA Purification Kit (ThermoFisher Scientific) according to the manufacturer’s instructions. Briefly, total RNA was incubated with the Dynabeads/binding buffer suspension at room temperature for 5 min and the reaction tubes were placed on a magnet until solution was clear. The supernatant containing poly(A)-RNA was saved. The beads with poly(A)+ RNA were washed three times with washing buffer provided by the kit. RNA was extracted from the supernatant and beads respectively using Trizol. qPCR was performed to analyze levels of FincoR and 36b4 in the Poly(A)+ and poly(A)-RNA fractions.

Genomics Analysis

RNA-Seq and GRO-Seq reads were mapped to the mouse reference genome mm10 with STAR aligner. Transcript quantifications were done with HOMER tool set and enhancer RNAs were identified based on H3K27ac ChIP-Seq in mouse liver. Briefly, intergenic H3K27ac ChIP-Seq peaks were selected as putative enhancer regions in mouse liver. Then +/-3kb regions around putative enhancers with RNA-Seq signal (>1 RPKM) were considered as putative eRNAs in mouse livers. Overlapped eRNAs regions were merged, and redundant ones were removed. In addition, any +/-3kb extended eRNAs regions that overlapped with protein-coding genes were further removed to avoid transcriptional readthrough from genes. We used DESeq2 to identify significantly regulated eRNAs with a cutoff of (FDR < 0.05, Log2FC > 1). Public ChIP-Seq datasets were obtained from ENCODE or GEO (Supplementary Table S3).

Statistical analysis

Statistical analysis was performed using GraphPad Prism 9. Statistical differences were evaluated using two-tailed unpaired Student’s t-test for comparisons between two groups, or ANOVA and appropriate post hoc analyses for comparisons of more than two groups. Statistical methods and corresponding p values for data shown in each panel were indicated in figure legends.

Data availability

RNA-seq and Gro-seq data were deposited in GEO under the following accession numbers GSE 221986.

eLife Assessment

The authors report a novel hepatic lncRNA FincoR regulating FXR with therapeutic implications in the treatment of MASH. The findings are important and use an appropriate methodology in line with the current state-of-the-art, with convincing support for the claims.

Reviewer #1 (Public Review):

Summary:

In the article titled "Hammerhead-type FXR agonists induce an eRNA FincoR that ameliorates nonalcoholic steatohepatitis in mice," the authors explore the role of the Farnesoid X Receptor (FXR) in treating metabolic disorders like NASH. They identify a new liver-specific long non-coding RNA (lncRNA), FincoR, regulated by FXR, notably induced by agonists such as tropifexor. The study shows that FincoR plays a significant role in enhancing the efficacy of tropifexor in mitigating liver fibrosis and inflammation associated with NASH, suggesting its potential as a novel therapeutic target. The study makes a promising contribution to understanding the role of FincoR in alleviating liver fibrosis in NASH, providing initial insights into the mechanisms involved. While it offers a valuable starting point, there is potential for further exploration into the functional roles of FincoR and their specific actions in human NASH cases. Building upon the current findings to elucidate more detailed mechanistic pathways through which FincoR exerts its therapeutic effects in liver disease would elevate the research's significance and potential impact in the field.

Strengths:

This study stands out for its comprehensive and unbiased approach to investigating the role of FincoR, a liver-specific lncRNA, in the treatment of NASH. Key strengths include: 1) The application of advanced sequencing methods like GRO-seq and RNA-seq offered a comprehensive and unbiased view of the transcriptional changes induced by tropifexor, particularly highlighting the role of FincoR. 2) Utilizing a genetic mouse model of FXR KO and a FincoR liver-specific knockdown (FincoR-LKD) mouse model provided a controlled and relevant environment for studying NASH, allowing for precise assessment of tropifexor's therapeutic effects. 3) The inclusion of tropifexor, an FDA-approved FXR agonist, adds significant clinical relevance to the study. It bridges the gap between experimental research and potential therapeutic application, providing a direct pathway for translating these findings into real-world clinical benefits for NASH patients. 4) The study's rigorous experimental design, incorporating both negative and positive controls, ensured that the results were specifically attributable to the action of FincoR and tropifexor.

Weaknesses:

The study presents several notable weaknesses that could be addressed to strengthen its findings and conclusions: 1) The authors focus on FincoR, but do not extensively test other lncRNAs identified in Figure 1A. A more comprehensive approach, such as rescue experiments with these lncRNAs, would provide a better understanding of whether similar roles are played by other lncRNAs in mitigating NASH. 2) FincoR was chosen for further study primarily because it is the most upregulated lncRNA induced by GW4064. Including another GW4064-induced lncRNA as a control in functional studies would strengthen the argument for FincoR's unique role in NASH. 3) The study does not conclusively demonstrate whether FincoR is specifically expressed in hepatocytes or other liver cell types. Conducting FincoR RNA-FISH with immunofluorescent experiments or RT-PCR, using markers for different liver cell types, would clarify its expression profile. 4) Understanding the absolute copy number of FincoR is crucial. Determining whether there are sufficient copies of FincoR to function as proposed would lend more credibility to its suggested role. 5) The manuscript, although technically proficient, does not thoroughly address the relevance of these findings to human NASH. Questions like the conservation of FincoR in humans and its potential role in human NASH should be discussed.

Reviewer #2 (Public Review):

Summary:

Nonalcoholic fatty liver disease (NASH), recently renamed as metabolic dysfunction-associated steatohepatitis (MASH) is a leading cause of liver-related death. Farnesoid X receptor (FXR) is a promising drug target for treating NASH and several drugs targeting FXR are under clinical investigation for their efficacy in treating NASH. The authors intended to address whether FXR mediates its hepatic protective effects through the regulation of lncRNAs, which would provide novel insights into the pharmacological targeting of FXR for NASH treatment. The authors went from an unbiased transcriptomics profiling to identify a novel enhancer-derived lncRNA FincoR enriched in the liver and showed that the knockdown of FincoR in a murine NASH model attenuated part of the effect of tropifexor, an FXR agonist, namely inflammation and fibrosis, but not steatosis. This study provides a framework for how one can investigate the role of noncoding genes in pharmacological intervention targeting known protein-coding genes. Given that many disease-associated genetic variants are located in the non-coding regions, this study, together with others, may provide useful information for improved and individualized treatment for metabolic disorders.

Strengths:

The study leverages both transcriptional profile and epigenetic signatures to identify the top candidate eRNA for further study. The subsequent biochemical characterization of FincoR using FXR-KO mice combined with Gro-seq and Luciferase reporter assays convincingly demonstrates this eRNA as a FXR transcriptional target sensitive to FXR agonists. The use of in vitro culture cells and the in vivo mouse model of NASH provide multi-level evaluation of the context-dependent importance of the FincoR downstream of FXR in the regulation of functions related to liver dysfunction.

Weaknesses:

As discussed, future work to dissect the mechanisms by which FincoR facilitates the action of FXR and its agonists is warranted. It would be helpful if the authors could base this on the current understanding of eRNA modes of action and the observed biochemical features of FincoR to speculate potential molecular mechanisms explaining the observed functional phenotype. It is unclear if this eRNA is conserved in humans in any way, which will provide relevance to human disease. Additionally, the eRNA knockdown was achieved by deletion of an upstream region of the eRNA transcription. A more direct approach to alter eRNA levels, e.g., overexpression of FincoR in the liver would provide important data to interpret its functional regulation.

Author Response

The following is the authors’ response to the original reviews.

eLife assessment:

The authors report a novel hepatic lncRNA FincoR regulating FXR with therapeutic implications in the treatment of MASH. The findings are important and use an appropriate methodology in line with the current state-of-the-art, with convincing support for the claims.

Public Reviews:

Reviewer #1 (Public Review):

Summary:

In the article titled "Hammerhead-type FXR agonists induce an eRNA FincoR that ameliorates nonalcoholic steatohepatitis in mice," the authors explore the role of the Farnesoid X Receptor (FXR) in treating metabolic disorders like NASH. They identify a new liver-specific long non-coding RNA (lncRNA), FincoR, regulated by FXR, notably induced by agonists such as tropifexor. The study shows that FincoR plays a significant role in enhancing the efficacy of tropifexor in mitigating liver fibrosis and inflammation associated with NASH, suggesting its potential as a novel therapeutic target. The study makes a promising contribution to understanding the role of FincoR in alleviating liver fibrosis in NASH, providing initial insights into the mechanisms involved. While it offers a valuable starting point, there is potential for further exploration into the functional roles of FincoR and their specific actions in human NASH cases. Building upon the current findings to elucidate more detailed mechanistic pathways through which FincoR exerts its therapeutic effects in liver disease would elevate the research's significance and potential impact in the field.

Strengths:

This study stands out for its comprehensive and unbiased approach to investigating the role of FincoR, a liver-specific lncRNA, in the treatment of NASH. Key strengths include: 1) The application of advanced sequencing methods like GRO-seq and RNA-seq offered a comprehensive and unbiased view of the transcriptional changes induced by tropifexor, particularly highlighting the role of FincoR. 2) Utilizing a genetic mouse model of FXR KO and a FincoR liver-specific knockdown (FincoR-LKD) mouse model provided a controlled and relevant environment for studying NASH, allowing for precise assessment of tropifexor's therapeutic effects. 3) The inclusion of tropifexor, an FDAapproved FXR agonist, adds significant clinical relevance to the study. It bridges the gap between experimental research and potential therapeutic application, providing a direct pathway for translating these findings into real-world clinical benefits for NASH patients. 4) The study's rigorous experimental design, incorporating both negative and positive controls, ensured that the results were specifically attributable to the action of FincoR and tropifexor.

Weaknesses:

The study presents several notable weaknesses that could be addressed to strengthen its findings and conclusions: 1) The authors focus on FincoR, but do not extensively test other lncRNAs identified in Figure 1A. A more comprehensive approach, such as rescue experiments with these lncRNAs, would provide a better understanding of whether similar roles are played by other lncRNAs in mitigating NASH. 2) FincoR was chosen for further study primarily because it is the most upregulated lncRNA induced by GW4064. Including another GW4064-induced lncRNA as a control in functional studies would strengthen the argument for FincoR's unique role in NASH. 3) The study does not conclusively demonstrate whether FincoR is specifically expressed in hepatocytes or other liver cell types. Conducting FincoR RNA-FISH with immunofluorescent experiments or RT-PCR, using markers for different liver cell types, would clarify its expression profile. 4) Understanding the absolute copy number of FincoR is crucial. Determining whether there are sufficient copies of FincoR to function as proposed would lend more credibility to its suggested role. 5) The manuscript, although technically proficient, does not thoroughly address the relevance of these findings to human NASH. Questions like the conservation of FincoR in humans and its potential role in human NASH should be discussed.

Reviewer #2 (Public Review):

Summary:

Nonalcoholic fatty liver disease (NASH), recently renamed as metabolic dysfunctionassociated steatohepatitis (MASH) is a leading cause of liver-related death. Farnesoid X receptor (FXR) is a promising drug target for treating NASH and several drugs targeting FXR are under clinical investigation for their efficacy in treating NASH. The authors intended to address whether FXR mediates its hepatic protective effects through the regulation of lncRNAs, which would provide novel insights into the pharmacological targeting of FXR for NASH treatment. The authors went from an unbiased transcriptomics profiling to identify a novel enhancer-derived lncRNA FincoR enriched in the liver and showed that the knockdown of FincoR in a murine NASH model attenuated part of the effect of tropifexor, an FXR agonist, namely inflammation and fibrosis, but not steatosis. This study provides a framework for how one can investigate the role of noncoding genes in pharmacological intervention targeting known protein-coding genes. Given that many disease-associated genetic variants are located in the non-coding regions, this study, together with others, may provide useful information for improved and individualized treatment for metabolic disorders.

Strengths:

The study leverages both transcriptional profile and epigenetic signatures to identify the top candidate eRNA for further study. The subsequent biochemical characterization of FincoR using FXR-KO mice combined with Gro-seq and Luciferase reporter assays convincingly demonstrates this eRNA as a FXR transcriptional target sensitive to FXR agonists. The use of in vitro culture cells and the in vivo mouse model of NASH provide multi-level evaluation of the context-dependent importance of the FincoR downstream of FXR in the regulation of functions related to liver dysfunction.

Weaknesses:

As discussed, future work to dissect the mechanisms by which FincoR facilitates the action of FXR and its agonists is warranted. It would be helpful if the authors could base this on the current understanding of eRNA modes of action and the observed biochemical features of FincoR to speculate potential molecular mechanisms explaining the observed functional phenotype. It is unclear if this eRNA is conserved in humans in any way, which will provide relevance to human disease. Additionally, the eRNA knockdown was achieved by deletion of an upstream region of the eRNA transcription. A more direct approach to alter eRNA levels, e.g., overexpression of FincoR in the liver would provide important data to interpret its functional regulation.

We thank the Editor and Reviewers for their constructive comments. We believe we have addressed all of the issues (detailed below) and the revisions have greatly strengthened the manuscript.

Reviewer 1:

The study presents several notable weaknesses that could be addressed to strengthen its findings and conclusions:

(1) The authors focus on FincoR, but do not extensively test other lncRNAs identified in Figure 1A. A more comprehensive approach, such as rescue experiments with these lncRNAs, would provide a better understanding of whether similar roles are played by other lncRNAs in mitigating NASH.

(2) FincoR was chosen for further study primarily because it is the most upregulated lncRNA induced by GW4064. Including another GW4064-induced lncRNA as a control in functional studies would strengthen the argument for FincoR's unique role in NASH.

(3) The study does not conclusively demonstrate whether FincoR is specifically expressed in hepatocytes or other liver cell types. Conducting FincoR RNA-FISH with immunofluorescent experiments or RT-PCR, using markers for different liver cell types, would clarify its expression profile.

(4) Understanding the absolute copy number of FincoR is crucial. Determining whether there are sufficient copies of FincoR to function as proposed would lend more credibility to its suggested role.

Response to 1 - 4): We thank Reviewer 1 for the positive comments on the strength of our work, including the open-ended approach, the novel eRNA FincoR and its strong relevance to liver disease. We also value the constructive feedback provided by the reviewer and agree that additional studies are important to fully understand the mechanisms of FincoR and the functional significance of other FXR-induced lncRNAs. In this manuscript we report the discovery and initial characterization of FincoR, as well as its potential function in FXR action in response to hammerhead agonists, but a number of interesting questions are raised. Future experiments, as suggested by reviewer, will be needed to examine the role of other FXR-induced lncRNAs, the potential role of FincoR induction by other nuclear receptors with binding sites at FincoR, whether FincoR is expressed in liver cell types in addition to hepatocytes, and the expression abundance of FincoR. These are all excellent suggestions for future experimentation which we feel are beyond the scope of the present report. For example, generating a genetic CRISPR/Cas9 of another lncRNA is not trial as it takes a significant amount of work with murine models. Also, we did not mean to exclude if other lncRNAs induced by FXR also bear functions. Technically, rescue experiment is not possible as FincoR RNA can be potentially very long (~10 kb if estimated by RNA-seq pattern in Fig.1C), and it is not feasible now to properly express it by exogenous vectors to ensure the expression levels are similar to endogenous ones. We therefore consider that these important questions are more suitable for future work to fully address. Our belief is that a comprehensive exploration of FXR-regulated lncRNAs holds the potential to unveil novel insights crucial for the development of therapies targeting NASH and other metabolic diseases. The study of FincoR is the beginning of this area of research.

(5) The manuscript, although technically proficient, does not thoroughly address the relevance of these findings to human NASH. Questions like the conservation of FincoR in humans and its potential role in human NASH should be discussed.

Response: These are important questions. To respond to the reviewer’s comment, new experiments are presented in our final revised manuscript in which we utilized mouse models of NAFLD/NASH and cholestatic liver injury to determine FincoR’s role in these diseases. Hepatic FincoR levels were significantly increased in mice fed with high fat diet (HFD) for 12 weeks (Supplementary Figure S1A) and in mice fed a HFD with high fructose (HFHF) in drinking water for 12 weeks (Supplementary Figure S1B). Elevated hepatic FincoR levels were also observed in mice treated with α-naphthylisothiocyanate (ANIT), a chemical inducer of liver cholestasis (Supplementary Figure S1C), and in mice with bile duct ligation (BDL), a surgical method to induce cholestatic liver injury (Supplementary Figure S1D).

In terms of the human relevance, we have provided additional information and figures showing that there is sequence similarity between mouse FincoR and a human loci. FincoR sequence is moderately conserved between mice and humans as displayed in the UCSC genome browser (Supplementary Figure S1E). Annotation of these conserved human sequences revealed that they overlap with a functionally uncharacterized human lncRNA XR_007061585.1 (Supplementary Figure S1F). Further, we conducted qRT-PCR experiment from human patient’s RNA samples, which demonstrated that hepatic lncRNA XR_007061585.1 levels are elevated in patients with NAFLD and PBC, but not in severe NASH-fibrosis patients (Supplementary Figure S1G, H). These results demonstrate that hepatic levels of a potential human analog of FincoR are elevated in NAFLD and PBC patients, which is consistent with FincoR’s upregulation in mouse models of chronic liver disease with hepatic inflammation and liver injury. Whether human lncRNA XR_007061585.1 is entirely analogous to mouse FincoR in terms of functions and mechanisms, and whether the elevation of this human lncRNA has a role in liver disease progression or is an adaptive response to liver injury remains to be determined.

Reviewer #2 (Recommendations For The Authors):

(1) In the introduction Line 96, "..., while the vast majority are transcribed into ncRNAs" may not be accurate. Please refer to Pointing and Haerty Annu Rev 2022 for a related discussion.

Response: We would like to thank the reviewer for pointing out this inaccurate information in the introduction. We have changed the content in the text, “While a significant portion of the genome was initially thought to be "junk DNA", it has been established that many non-coding regions give rise to functional non-coding RNAs.”

(2) Figure 5: the authors should provide a clear illustration demonstrating the sequence targeted by the sgRNA in relation to the transcriptional and epigenetic profile (i.e., RNAseq and H3K27ac ChIP-seq data).

Response: The illustration (Figure 5-figure supplement 1A, right panel) demonstrating the sequence targeted by the sgRNA has been updated as suggested by the reviewer.

In this model, the upstream of FincoR is deleted, leading to the inhibition of FincoR transcription. Does the deleted region include FXR binding sites? If so, would the phenotype be due to the deletion of these binding sequences, rather than the decreased FincoR transcripts? Accordingly, the limitation or alternative interpretation should be discussed.

Response: The reviewer made a good point. The deleted region includes FXR binding sites so that we cannot rule out decreased binding of FXR or decreased transcription of the region per se, in addition to the decreased levels of FincoR, to bear a role in the phenotypic changes we observed. In the final revision, we have added discussion of this alternative (6th paragraph in the revised discussion section).

(3) Figure 6C, the images should be accompanied by quantification. It appears the FincoR-KD shows a visible difference as compared to Tropifexor-treated control mice, which does not match entirely what is written in the results.

Response: The quantitation of Oil Red O staining has been done as suggested by the reviewer (Figure 6C). The result is consistent with the triglyceride result showing that tropifexor treatment markedly reduced neutral lipids determined by Oil Red O staining of liver sections (Figure 6C) and liver TG levels (Figure 6D) and these beneficial effects on reducing fatty liver were not altered by FincoR.

(4) Figure 7, does AST show the same pattern as ALT? As indicated from Line 335, "tropifexor treatment reduced mRNA levels of several genes that promote fibrosis (Col1a1, Col1a2, ...)". Fig. 7D does not seem to match the description of Col1a1. Authors may need to modify the results.

Response: AST has been measured and has the same pattern as ALT. The new data have been added to Figure 7B. Col1a1 expression has been re-measured and the results have been updated in Figure 7D.

(5) Is FincoR level reduced in NASH conditions?

Response: We thank the Reviewer for this question. We now added new data to examine the levels of FincoR in mouse liver disease models and also examined levels of a potential human analog of FincoR in human liver specimens from PBC, NAFLD, and NASH patients. Please see our new data and description above in the response to comment 5 by Reviewer 1 (most data now included in the new Supplementary Figure S1).

(6) Please provide information on the conservation of FincoR (DNA and RNA) in humans. This would be important to provide the human disease relevance.

Response: As described above in the response to comment 5 of reviewer 1, a human loci shows sequence similarity to mouse FincoR and this conserved region has an annotated uncharacterized human lncRNA. We also examined the levels of this human homolog in human diseased liver samples. Our new results demonstrate that hepatic levels of a potential human analog of FincoR are elevated in NAFLD and PBC patients, which is consistent with FincoR’s upregulation in mouse models of chronic liver disease with hepatic inflammation and liver injury. Whether human lncRNA XR_007061585.1 is entirely analogous to mouse FincoR in terms of functions and mechanisms, and whether the elevation of this human lncRNA has a role in liver disease progression or is an adaptive response to liver injury remains to be determined.

(7) Several discussion points for the authors' consideration:

(7.1) human-mouse conservation as alluded to in #6;

Response: Potential human-mouse conservation is discussed with new data in the last paragraph of the Results section.

(7.2) potential molecular mechanism involved in FincoR-regulated hepatocyte function;

Response: We thank Reviewer for this comment. We have added more discussion as shown below: “RNA inside the cells usually associates with different RNA-binding proteins (RBPs). To predict those potential binding proteins of FincoR. Additional bioinformatic analysis identified proteins that potentially binding FincoR, including KHDRBS1, RBM38, YBX2 and YBX3 (Supplemental Table S5). These findings and potential functions of the binding proteins are discussed in the 5th paragraph of the discussion section in the final revised manuscript. Whether these predicted RBPs interact with FincoR and the underlying mechanisms will need to be investigated in future experimentation to understand the mechanisms involved in FincoR-regulated hepatocyte function.”

(7.3) any disease-associated SNPs in the FincoR locus.

Response: No SNPs were noted in the annotation of the human loci with sequence similarity to mouse FincoR in the NCBI genome data viewer.

(7.4) the in vitro induction of FincoR is transient but in vivo this occurs after 12 days of drug treatment. How do the authors reconcile the differential induction patterns?

Response: To clarify, the induction of FincoR after a single dose of GW4064 in vivo was transient, peaked within 1 h and then declined gradually (Figure 1-figure Supplement 1C). In the tropifexor treatment protocol (also in vivo), the mice were treated daily with tropifexor for 12 days so that the multiple doses maintained FincoR induction. The beneficial effect of tropifexor by inducing FincoR, therefore, accumulated over the 12 days.

It is worthy to note that we failed to see induction of FincoR in isolated primary mouse hepatocytes treated with GW4064 in vitro. We can only detect FincoR in primary hepatocytes isolated from GW4064-treated mice liver. This may be due to the loss of key factors mediating FincoR induction in the cultured primary hepatocytes.

Version 2

Abstract

Abstract

The nuclear receptor, Farnesoid X Receptor (FXR/NR1H4), is increasingly recognized as a promising drug target for metabolic diseases, including nonalcoholic steatohepatitis (NASH). Protein coding genes regulated by FXR are well known, but whether FXR also acts through regulation of long non-coding RNAs (lncRNAs), which vastly outnumber protein-coding genes, remains unknown. Utilizing RNA-seq and GRO-seq analyses in mouse liver, we found that FXR activation affects the expression of many RNA transcripts from chromatin regions bearing enhancer features. Among these we discovered a previously unannotated liver-enriched enhancer-derived lncRNA (eRNA), termed FincoR. We show that FincoR is specifically induced by the hammerhead-type FXR agonists, including GW4064 and tropifexor. CRISPR/Cas9-mediated liver-specific knockdown of FincoR in dietary NASH mice reduced the beneficial effects of tropifexor, an FXR agonist currently in clinical trials for NASH and primary biliary cholangitis (PBC), indicating that that amelioration of liver fibrosis and inflammation in NASH treatment by tropifexor is mediated in part by FincoR. Overall, our findings highlight that pharmacological activation of FXR by hammerhead-type agonists induces a novel eRNA, FincoR, contributing to the amelioration of NASH in mice. FincoR may represent a new drug target for addressing metabolic disorders, including NASH.

Introduction

Nonalcoholic fatty liver disease (NAFLD) is the most common chronic liver disease and a leading cause of liver transplants and liver-related death (Friedman et al., 2018). NAFLD begins with simple steatosis but may further progress to a severe form, non-alcoholic steatohepatitis (NASH), and later, fatal cirrhosis and liver cancer (Friedman et al., 2018). Despite its striking global increase and clinical importance, there is no approved drug for NASH. The urgent need for development of therapeutic agents for NASH has greatly increased research interest in the nuclear receptor, Farnesoid X Receptor (FXR, NR1H4) (Evans & Mangelsdorf, 2014).

FXR is activated by its physiological ligands, bile acids (BAs), and regulates expression of genes involved in BA, lipid, and glucose metabolism and hepatic autophagy, which maintain metabolite levels and metabolic homeostasis (Calkin & Tontonoz, 2012; Kliewer & Mangelsdorf, 2015; Lee et al., 2006; Lee et al., 2014; Seok et al., 2014). Ligand-activated FXR also protects against hepatic inflammation and liver injury (Jung et al., 2020; Wang et al., 2008). The action of FXR, similar to other nuclear receptors, is achieved primarily by its binding to chromatin loci to regulate the transcription of target genes (Calkin & Tontonoz, 2012; Lee et al., 2006; Lee et al., 2012; Thomas et al., 2010). Consistent with its crucial physiological functions, FXR is increasingly recognized as a promising drug target, particularly for liver diseases, such as NASH and primary biliary cholangitis (PBC) (Abenavoli et al., 2018; Ali et al., 2015; Downes et al., 2003; Kremoser, 2021). For example, semi-synthetic or non-steroidal synthetic agonists of FXR, including obeticholic acid (OCA) and hammerhead-type agonists, such as tropifexor and cilofexor, are currently in clinical trials for NASH and PBC patients (Abenavoli et al., 2018; Kremoser, 2021; Sanyal et al., 2023; Tully et al., 2017). However, how pharmacological activation of FXR mediates such beneficial therapeutic effects is poorly understood.

Non-protein-coding RNAs (ncRNAs) are one of the fascinating discoveries of modern biology (Cech & Steitz, 2014). While a significant portion of the genome was initially thought to be “junk DNA”, it has been established that many non-coding regions give rise to functional non-coding RNAs (Cech & Steitz, 2014). Of these ncRNAs, long non-coding RNAs (lncRNAs) are a group of transcripts longer than 200 nucleotides and play important roles in diverse biological processes (Lam et al., 2014; Li et al., 2016; Mattick et al., 2023; Sartorelli & Lauberth, 2020; Statello et al., 2021). A group of lncRNAs are produced from genomic regions bearing epigenetic features of enhancers (Hon et al., 2017; Mattick et al., 2023). This is consistent with the idea that many transcriptional enhancers actively transcribe noncoding RNAs that are referred to as eRNAs (>85k in humans and >57k in mice) (Hirabayashi et al., 2019), some of which are functionally important for enhancer functions (Lai & Shiekhattar, 2014; Li et al., 2016; Sartorelli & Lauberth, 2020). A majority of eRNAs have not yet been annotated in current lncRNA databases, such as GENCODE. Exploring eRNA landscapes and functions in diverse biology and disease states will facilitate our understanding of both lncRNAs and enhancers (Li et al., 2016; Mattick et al., 2023). Indeed, the landscape of eRNAs in mouse liver has been minimally explored (Fang et al., 2014), and has not been well-studied in response to specific nuclear receptor activation, such as by FXR. Moreover, functional roles of eRNAs in vivo in intact organisms is understudied.

FXR was shown to regulate expression of small non-coding microRNA (miR) genes, for example miR-34a and miR-802 (Lee et al., 2010; Seok et al., 2021), but it has not been reported whether FXR achieves its function through regulating lncRNAs, which far outnumber miRs. Because FXR directly regulates expression of its target genes (Lee et al., 2012; Thomas et al., 2010), we examined if FXR regulates enhancer-derived lncRNAs (eRNAs), and if such transcripts participate in its physiological or pharmacological functions. By utilizing RNA-seq and GRO-seq analyses of livers from mice treated with FXR ligands, we identified a set of FXR-regulated eRNAs. Among these, we focused on a highly induced and abundantly expressed eRNA that we referred to as FXR-induced non-coding RNA (FincoR) for functional studies. FincoR is highly enriched in mouse liver and is induced specifically by hammerhead-type FXR agonists, including GW4064 and tropifexor. In vivo studies utilizing CRISPR/Cas9-mediated liver-specific knockdown of FincoR in dietary NASH mice indicated that FincoR is critically involved in mediating the beneficial pharmacological effects of tropifexor in reducing liver fibrosis and inflammation.

Results

Activation of FXR by GW4064 induces a novel eRNA, FincoR, in mouse liver

To identify enhancer-derived lncRNAs (eRNAs) potentially regulated by FXR, we first obtained a list of putative enhancers in mouse liver based on ENCODE H3K27ac ChIP-Seq data (ENCFF001KMI, see methods). Then, we performed ribo-depleted total RNA-seq in the livers from mice treated with a specific FXR agonist, GW4064, to identify transcripts produced from these enhancer regions (see Supplementary Table S1). To avoid confounding issues of RNA signals from genes, we specifically focused on intergenic enhancer regions (+/- 3kb from H3K27ac peak center) that harbor discernible RNA-Seq signals (RPKM > 1).

This genomic analysis resulted in identification of 190 high-confidence eRNAs in mouse liver. Among these, 14 eRNAs were upregulated and 5 were downregulated by GW4064 treatment (FDR < 0.05, Log2 FC > 1, Figure 1A) (see Supplementary Table S2). FXR-regulated eRNAs were produced adjacent to many genes with important roles in liver metabolism and disease, for example, Hes1 (Figure 1-figure supplement 1A, B). One of the most robustly induced eRNAs was FXR-induced non-coding RNA (FincoR), an unannotated novel transcript located on chromosome 19 (Figure 1A, B, C). Reverse transcription qPCR (RT-qPCR) confirmed that treatment with GW4064 substantially induced expression of FincoR, more than 10-fold in mouse liver, which is similar to induction of Shp, a well-known FXR target gene (Figure 1D) (Claudel et al., 2005; Evans & Mangelsdorf, 2014; Goodwin et al., 2000). Induction of FincoR by GW4064 was transient, peaked within 1 h and then declined gradually, a pattern similar to that of Shp (Figure 1-figure supplement 1C). Expression of genes adjacent to FincoR, including Gcnt1, Rfk, Pcsk5 and Prune2, did not change after acute 1 h FXR activation as shown by RNA-seq (Figure 1C), and confirmed for Gcnt1 by time course qPCR (Figure 1-figure supplement 1C).

Activation of FXR by GW4064 induces FincoR, a novel eRNA, in mouse liver.

(A) Volcano plot from RNA-Seq showing significantly induced eRNAs (FincoR is highlighed) in the livers of C57BL/6 male mice treated with GW4064 (i.p. injection, 30 mg/kg, 1 h) or vehicle. The x axis denotes log2 fold change (GW4064/Veh) of eRNAs and the y axis denotes -log10 FDR of eRNAs. (B) A Bar plot showing the reads per million (RPMs) of up regulated eRNAs by GW4064 treatment. (C) IGV genome browser track showing RNA-seq signals from vehicle or GW4064 treated samples around the FincoR locus and its neighboring regions. A zoom-in view of FincoR is shown below. Veh, vehicle; GW, GW4064; pos, positive strand; neg, negative strand; Refseq, reference sequence. (D) RT-qPCR data showing GW4064 induction of FincoR eRNA and Shp mRNA in the liver (n=4/group). Data are presented as mean ± SEM. Statistical significance was determined by the two-way ANOVA Sidak’s multiple comparisons test with **p < 0.01 and ***p < 0.001.

We also found that the short-time GW4064 treatment resulted in 590 up-regulated genes and 500 down-regulated genes (FDR < 0.05) (Figure 1-figure supplement 2A). GO enrichment analysis of these differentially expressed genes (DEGs) revealed their roles in the regulation of triglyceride, fatty acid, and cholesterol metabolism (Figure 1-figure supplement 2B), which is consistent with known roles of FXR in these physiological processes (Claudel et al., 2005).

Ligand-activated FXR directly activates transcription of eRNAs, including FincoR

We sought to examine, 1) if these eRNAs were directly activated by FXR, and 2) if the activation takes place transcriptionally using FXR liver-specific knockout (FXR-LKO) mice that were treated with GW4064 to determine if FXR was required for induction of eRNAs by GW4064. FXR-LKO mice were generated from FXR floxed mice (FXR-flox) (Figure 2-figure supplement 1A) and transcription of eRNAs was detected by global run-on sequencing (GRO-seq), a widely used method to detect nascent RNA transcription, including eRNAs (Lam et al., 2014; Li et al., 2016; Sartorelli & Lauberth, 2020). GRO-seq showed that FXR-induced eRNAs were activated transcriptionally by GW4064 in FXR- flox mice, but such induction was abolished in FXR-LKO mice (Figure 2A). In particular, FincoR is robustly induced in the GRO-Seq analysis and its induction is dependent on hepatic FXR (Figure 2B). RT-qPCR confirmed the FXR-dependent expression of FincoR, similar to that of Shp (Figure 2-figure supplement 1B).

Ligand-activated FXR directly activates transcription of eRNAs including FincoR in the liver.

(A) Left: experimental outline. FXR-Flox and LKO male mice were fasted overnight and treated with vehicle or GW4064, and livers were collected 1 h later with nuclei isolated for Gro-seq (n=2/group). Right: a boxplot shows the Gro-Seq signals for GW4064 up-regulated eRNAs in different conditions. RPKM: reads per kbp per million. (B) IGV genome browser track showing RNA-seq, Gro-seq and ChIP-seq signals in the FincoR locus. An arrow at the bottom points to the main enhancer FXR ChIP-seq peak that overlaps the start site of FincoR. (C) ChIP assays were performed in the same liver samples described in Fig. 1A to detect FXR, RXR𝛂, and BRD4 occupancy at the FXR binding peak region close to the transcription start site of FincoR (black arrow in Fig.2B). (D) HepG2 cells were transfected with luciferase reporter expressing wild type FXRE or mutant FXRE (see materials and methods) for 24 h before treatment with GW4064 for an additional 6 h. Relative luciferase activities are shown. (C-D) Data are presented as mean ± SEM. Statistical significance was determined by the Student’s t test with *p < 0.05 and ***p < 0.001.

We next examined if FXR binds to the enhancers that produce the identified eRNAs by analyzing published mouse liver ChIP-Seq data for FXR (Supplementary Table S3) (Lee et al., 2012; Thomas et al., 2010). FXR binding was strongly enriched at the enhancers associated with FXR-induced eRNAs as were the enhancer marks H3K27ac and H3K4me1 (Figure 2-figure supplement 1C). The binding of FXR and the presence of histone marks at the FincoR enhancer region determined by ChIP-seq as compared with nascent transcripts detected by Gro-seq is shown in Figure 2B. These analyses support the conclusion that activation of FXR transcriptionally induces this series of eRNAs via chromatin binding at these enhancers, including FincoR.

We validated FXR binding at the enhancer region that produces FincoR using mouse liver ChIP (Figure 2C). We also examined binding of a well-known DNA binding partner of FXR, Retinoid x Receptor alpha (RXRα/NR2B1) (Evans & Mangelsdorf, 2014; Zheng et al., 2018), and bromodomain-containing protein 4 (BRD4), an acetylated histone reader protein that often binds at active enhancers (Chen et al., 2016; Li et al., 2016; Rahnamoun et al., 2018; Sartorelli & Lauberth, 2020) and is a transcriptional coactivator of FXR (Jung et al., 2020). GW4064 treatment resulted in substantial increases in recruitment of both FXR and RXRα to the enhancer region close to the transcription start site of FincoR (arrow shown below in Figure 2B), whereas binding was not detected at a control region (Figure 2C). We also found that BRD4 occupancy was increased at this enhancer region after GW4064 treatment (Figure 2C). These results indicate that GW4064 activation of FXR leads to increased occupancy of the FXR/RXRα heterodimer and BRD4 to the enhancer region to upregulate FincoR eRNA in the liver.

We identified an inverted repeat1 (IR1) motif that is known to bind FXR (Calkin & Tontonoz, 2012; Lee et al., 2006) within the major FXR binding peak near the start site of FincoR (arrow shown below in Figure 2B), which we also refer to as FXRE (Figure 2B, C). We examined the functionality of this IR1 motif for mediating transcriptional activation by GW4064 using reporter assays (Figure 2D). We cloned the region containing the IR1 motif into the pGL4.23 luciferase reporter and generated a mutated IR1 motif construct as a comparison (Figure 2D). After transfection into human hepatic HepG2 cells, GW4064 treatment significantly elevated the luciferase activity of the reporter with the wild type IR1 motif, but not with the mutated IR1 motif (Figure 2D). Together, these results suggest that GW4064-activated FXR directly upregulates FincoR expression.

FincoR is a liver-specific nucleus-enriched eRNA

Because enhancers and eRNAs generally act in a tissue-specific manner (Li et al., 2016; Sartorelli & Lauberth, 2020), we examined the tissue-specific expression of FincoR in mice. Strikingly, FincoR is highly expressed in the liver and it is expressed at extremely low levels in most other tissues, except for a detectable, but still fairly low, level in the lung (Figure 3A). GW4064 treatment resulted in induction of FincoR specifically in the liver (Figure 3A). The level of FincoR detected in primary mouse hepatocytes (PMHs) isolated from GW4064-treated mouse liver was similar to that in the liver tissue from the same mouse, suggesting that the majority of FincoR is present in hepatocytes (Figure 3B).

FincoR is a liver-specific nucleus-enriched eRNA.

(A) Expression levels of FincoR in various tissues after GW4064 treatment. Data from C57BL/6 male mice fasted overnight and i.p. injected with vehicle or GW4064 (30 mg/kg) for 1 h (n=2/group). (B) C57BL/6 male mice were fasted overnight and injected i.p. with vehicle or GW4064 (30 mg/kg) for 1 h. One small piece of liver was snap-frozen for later RNA isolation and the remaining part was used for immediate primary hepatocyte isolation. Then, RNAs were extracted from liver or primary hepatocytes and FincoR expression was measured (n=3/group). (C) Agarose gel electrophoresis of PCR products generated in 5′ (left) and 3′ (right) RACE of FincoR in liver samples. Primer locations are shown. RACE, rapid amplification of cDNA ends. GSP: gene specific primer. (D) A schematic diagram showing location of FincoR relative to nearby genes in the mice genome. (E) PhyloCSF analysis of the coding potential of FincoR. (F) In vitro translation of FincoR using the Promega Transcend Non-Radioactive Translation Detection Systems. Luciferase is used as a control for coding RNA. (G) qPCR analysis of FincoR, 36b4 in Poly(A)+ and Poly(A)-RNA fractions from GW4064 treated mouse liver. (H) FincoR identified in the subcellular fractions using cellular fractionation assays. Primary hepatocytes were isolated from GW4064 or DMSO treated mice and the cytoplasm and nucleus fractions of these hepatocytes were separated and both fractions were subjected to RNA extraction and qPCR.

By using the 5’ and 3’ rapid amplification of cDNA ends (RACE), we identified one transcript of FincoR that is of approximately 3.7 kb in length (Figure 3C). However, based on RNA-seq, the length of FincoR is over 10 kb (Figure 1C), suggesting there are likely additional multiple RNA isoforms that we were not able to identify by RACE. We next analyzed the coding potential of FincoR utilizing a comparative genomic program, PhyloCSF. While adjacent genes Gcnt1 and Prune2 (Figure 3D, E) were correctly predicted to encode proteins, FincoR did not contain a potential protein-coding open reading frame (Figure 3E). Consistent with this bioinformatic prediction, a vector expressing FincoR failed to produce any proteins in an in vitro transcription/translation assay (Figure 3F) confirming that the FincoR transcript is a noncoding RNA. FincoR transcripts were enriched by binding to oligo-dT beads, suggesting that FincoR is 3’ polyadenylated (Figure 3G). Further, FincoR was detected in the nuclear compartment and GW4064 treatment increased the nuclear abundance of FincoR (Figure 3H), consistent with its potential transcriptional regulatory function. Together, these results from molecular biochemical characterization studies reveal that FincoR is a liver-enriched nuclear polyadenylated eRNA.

FincoR is induced specifically by the hammerhead-type synthetic FXR agonists

To determine whether induction of the FXR-induced FincoR is ligand-specific, we examined the effects of several hammerhead-type synthetic FXR agonists, including GW4064, cilofexor, and tropifexor; a semi-synthetic agonist, OCA; and a non-hammerhead-type gut-specific agonist, fexaramine (Figure 4A) (Downes et al., 2003; Fang et al., 2015). Remarkably, treatment with each of the hammerhead type agonists for 1 h resulted in a robust induction of FincoR (Figure 4B), whereas FincoR levels were unchanged after treatment with OCA or fexaramine for 1 h (Figure 4B). Treatment with OCA for 4 h or even one-week treatment with OCA failed to induce hepatic FincoR in mice, while expression of Shp was significantly induced (Figure 4C, D). Acute feeding with a diet supplemented with 0.5% cholic acid (CA), a primary BA, for 6 h also failed to induce FincoR expression (Figure 4E). Collectively, these results demonstrate that FincoR is induced specifically by hammerhead-type FXR agonists.

FincoR is induced by the hammerhead class of non-steroidal FXR agonists, including GW4064 and tropifexor.

(A) The chemical structures of the FXR agonists including the hammerhead class of synthetic FXR agonists, non-hammerhead-type synthetic agonists, semi-synthetic BA and natural BA. BA: bile acid. (B) qPCR data showing FincoR expression levels in C57BL/6 mice liver respectively treated with GW4064 (30 mg/kg), Cilofexor (30 mg/kg), Tropifexor (0.5 mg/kg), Fexaramine (100 mg/kg), or OCA (20 mg/kg) for 1 h (n=3∼4/group). (C) FincoR expression levels in C57BL/6 mice liver treated with OCA (20 mg/kg) or Fexaramine (100 mg/kg) for 4 hrs (n=3/group). (D) Expression of FincoR in C57BL/6 mice liver after daily treatment with OCA (20 mg/kg) for 7 days (n=5/group). Shp gene mRNA was measured as a positive control. (E) Expression of FincoR in C57BL/6 mice fed with 0.5% cholic acid (CA) diet for 6 hrs (n=3∼4/group). In panels B-E, all mice underwent over-night fasting. (B-E) Data are presented as mean ± SEM. Statistical significance was determined by the Student’s t test with *p < 0.05, **p < 0.01 and ***p < 0.001.

Generation of CRISPR/Cas9-mediated FincoR liver-specific knockdown mice

To explore the functional role of hepatic FincoR, we utilized the CRISPR/Cas9 technique to generate FincoR liver-specific knockdown (FincoR-LKD) mice (Figure 5-figure supplement 1A, B). Adenoviral-mediated sgRNA expression in Cas9 mice resulted in downregulation of FincoR specifically in the liver by about 60%, but not in other tissues (Figure 5B). As FXR is a key regulator of bile acid, cholesterol, lipid, and glucose metabolism and FincoR is specifically regulated by FXR, FincoR may have a role in the metabolic process in physiology and disease. We examined liver triglyceride, cholesterol, bile acid, glycogen and serum non-esterified fatty acids (NEFA) but FincoR downregulation did not result in any significant changes under physiological conditions (Figure 5C).

Generation of CRISPR/Cas9-mediated FincoR liver-specific knockdown mice.

(A) Experimental scheme: male Cas9 mice were infected with adenovirus expressing sgRNA for FincoR or a control for 1 week. Then the liver and serum were collected from these mice after 4∼5 hrs of fasting. (B) The expression of FincoR in the liver was measured by qPCR (n=4/group). (C) Hepatic triglyceride, cholesterol, bile acid, glycogen and serum NEFA were measured (n=5∼7/group) (D) Male Cas9 mice were infected with adenovirus expressing sgRNA for FincoR or control for 1 week. Then these mice were fasted overnight and treated with GW4064 for 3 h before tissue collection. RNA-seq profiles of expression of hepatic FincoR and the adjacent genes were shown (n=2/group). (E) Genome-wide changes in mRNA expression shown in a volcano plot. The numbers refer to the number of genes up- or down-regulated by 2-fold or more with an adjusted p-value < 0.01. (F) Gene ontology analysis of biological pathways using DAVID Tools for genes downregulated after FincoR knockdown. (B) Data are presented as mean ± SEM. Statistical significance was determined by the Student’s t test with ***p < 0.001.

To explore the molecular signatures and pathways affected by FincoR, we examined global gene expression by RNA-seq analysis in mouse liver after FincoR knock-down (Figure 5D). While FincoR was markedly downregulated, the neighboring genes, such as Gcnt1, Rfk, Pcsk5 and Prune2, were largely unchanged (Figure 5D; Figure 5-figure supplement 1C). The RNA-seq analysis revealed 18 up-regulated genes and 53 downregulated genes in FincoR-LKD liver (Supplementary Table S4; Figure 5E). GO analysis of those downregulated genes indicated that these genes were enriched in pathways involved in fatty acid oxidation, organelle organization and metabolic process (Figure 5F). Among the down-regulated genes, Ppp1r3g, which has a role in controlling glycogen synthesis, and Igfbp2, which functions in insulin resistance, were markedly reduced (Supplementary Table S4; Figure 5-figure supplement 1D, E). Among the up-regulated genes, expression was substantially increased for Eda2r, which is a member of the tumor necrosis factor receptor superfamily and is involved in inflammation, the immune response, and development (Figure 5-figure supplement 1F). Fndc1, which is involved in fibronectin matrix remodeling, was also suppressed by FincoR (and thus upregulated upon its KO, Figure 5-figure supplement 1G). These studies suggest that FincoR has a role in modulating metabolic homeostasis by regulating genes involved in metabolism and inflammation.

Amelioration of hepatic steatosis mediated by tropifexor is independent of FincoR in diet-induced NASH mice

Tropifexor, also known as LJN452, is a highly potent hammerhead type FXR agonist that is currently under clinical trials for NASH and PBC patients (Kremoser, 2021; Sanyal et al., 2023; Tully et al., 2017). Because FincoR is induced specifically by the hammerhead class of FXR agonist (Figure 4) and has a potential role in the regulation of metabolism and inflammation (Figure 5), we hypothesized that FincoR may play a role in tropifexor-mediated beneficial effects on reducing NASH pathologies in mice.

We utilized a mouse model that had been fed the Amylin Liver NASH-promoting (AMLN) diet (Hernandez et al., 2019; Sun et al., 2022; Zhao et al., 2018), and examined the potential impact of liver-specific downregulation of FincoR on tropifexor’s effects on NASH pathology. Cas9 mice fed the AMLN diet for 12 weeks were injected via tail veins with adenovirus expressing control sgRNA or FincoR sgRNA, respectively, and then treated daily with tropifexor (0.3 mg/kg) for 12 days (Figure 6A). In these mice, as a technical validation of RNA induction and knockdown, FincoR levels were significantly increased by FXR agonist tropifexor, and the increase was blocked by adenovirus expressing sgRNA for FincoR (Figure 6B).

In diet-induced NASH mice, tropifexor-mediated beneficial effects on reducing hepatic steatosis are largely independent of FincoR.

(A-E) Male Cas9 mice were fed a NASH diet for 12 weeks. The mice were randomly assigned to 3 groups and infected with adenovirus expressing sgRNA for FincoR or control. Three days later, the mice were treated with tropifexor (0.3 mg/kg) or vehicle for 12 days. The mice were administered tropifexor or vehicle and fasted for 4 h before tissues were collected. (A) Experimental scheme. (B) Hepatic FincoR expression was measured (n=6∼7/group). (C) Oil Red O staining of liver sections. Scale bar (50 μm). Image analyses were done using Image J and the areas of stained field were quantified (n=5/group) (D) Hepatic TG, hepatic cholesterol, gallbladder BA and hepatic BA levels were measured (n=6∼7/group). (E) mRNA levels in the liver of the indicated genes involved in bile acid regulation and lipid regulation (n=6∼7/group). (B, D and E) Data are presented as mean ± SEM. Statistical significance was determined by the one-way ANOVA (Sidak’s multiple comparisons test) with *p < 0.05, **p < 0.01 and ***p < 0.001. Ad, adenovirus; H & E, hematoxylin and eosin; TG, triglyceride; Veh, vehicle; ns, not significant.

We then examined the effect of tropifexor treatment and FincoR downregulation on hepatic steatosis in these mice. Tropifexor treatment markedly reduced neutral lipids determined by Oil Red O staining of liver sections (Figure 6C) and liver TG levels (Figure 6D), and these beneficial effects on reducing fatty liver were not altered by FincoR downregulation. Also, FincoR downregulation had little effect on liver cholesterol and gallbladder BA levels, although gallbladder BA levels were reduced by tropifexor (Figure 6D).

Consistent with the phenotypes, hepatic expression of key genes involved in BA synthesis was dramatically reduced by tropifexor treatment (i.e., Cyp7a1 and Cyp8b1), which is consistent with decreased gallbladder BA levels mediated by this agonist (Figure 6E). Similarly, tropifexor also lowered lipid synthesis genes (Srebp1c, Lpin1, Scd1), consistent with decreased liver TG levels. However, downregulation of FincoR did not result in changes in mRNA levels of these genes (Figure 6E). These results indicate that tropifexor-mediated beneficial effects on reducing hepatic steatosis are independent of FincoR.

FincoR facilitates alleviation of liver inflammation by tropifexor in diet-induced NASH

Tropifexor ameliorated fibrotic NASH pathologies in preclinical studies (Hernandez et al., 2019; Tully et al., 2017) and has recently concluded phase 2 clinical trials for NASH patients (Sanyal et al., 2023). We, therefore, further examined the effects of FincoR downregulation on altering other NASH pathologies, including hepatocellular apoptosis, liver fibrosis and inflammation.

In the same AMLN diet-fed mice as described above (Figure 6), analyses of liver sections revealed that tropifexor treatment reduced hepatocyte swelling/ballooning (H&E staining), decreased numbers of apoptotic cells (TUNEL staining), alleviated fibrosis (Sirius red staining), and lowered infiltration of macrophages (F4/80 staining) (Figure 7A). Remarkably, these tropifexor-mediated beneficial effects on NASH pathologies were all markedly diminished by FincoR downregulation (Figure 7A). This was consistently found in various liver lobes and two representative pictures from two different lobes are shown (Figure 7A). In control experiments, FincoR downregulation in vehicle-treated mice did not result in marked changes in NASH pathologies (Figure 7-figure supplement 1). These results demonstrate that FincoR is required for tropifexor-mediated beneficial effects on reducing NASH pathologies, specifically reducing hepatic inflammation, fibrosis and hepatocyte apoptosis. Consistent with these results, serum ALT and AST levels, indicators of liver damage, were significantly elevated after FincoR downregulation (Figure 7B). Protein levels of key inflammatory markers, IL1β and CCL2, in liver extracts were also elevated after FincoR downregulation (Figure 7C).

In diet-induced NASH mice, tropifexor-mediated beneficial effects on reducing liver fibrosis and inflammation are diminished by FincoR downregulation.

(A) Representative images from H & E, F4/80, Sirius Red and TUNEL staining of liver sections from the same cohort of mice described in Figure 6. Scale bar (50 μm). Image analyses were done using Image J and the area of collagen staining, TUNEL and F4/80 levels were quantified (n=5/group). (B) Serum ALT and AST levels were measured (n=5/group). (C) IL-1β and CCL2 levels in the liver tissues were determined by ELISA (n=5/group). (D) mRNA levels in the liver of the indicated genes involved in inflammation, fibrosis and cell death (n=5/group). (E) Model: FincoR is a liver-enriched eRNA that is induced specifically by hammerhead-type FXR agonists (top). In diet-induced NASH mice, FincoR is required for tropifexor-mediated beneficial effects on reducing hepatic inflammation, fibrosis, and cell death with the mechanisms to be determined (bottom). (A-D) Data are presented as mean ± SEM. Statistical significance was determined by the one-way ANOVA (Sidak’s multiple comparisons test) with *p < 0.05, **p < 0.01 and ***p < 0.001.

Consistent with the phenotypes from histological analyses, hepatic expression of fibrosis and inflammation was altered by FincoR knockdown. For example, tropifexor treatment reduced mRNA levels of several genes that promote fibrosis (Col1a1, Col1a2, Acta2) and hepatic inflammation (Eda2r, Ifng, Ccl3), whereas these reductions were largely reversed by FincoR downregulation (Figure 7D). We also detected increased expression of inflammatory genes (Ccl2, Ccr2, Lcn2) and an extracellular matrix remodeling gene (Fndc1) in liver with FincoR knockdown (Figure 7D). Tropifexor treatment suppresses hepatic apoptosis by reducing pro-apoptotic genes (Ctsb, Ctss) and upregulating anti-apoptotic genes such as Bcl2 (Warren et al., 2019). Importantly, these effects were significantly reversed by FincoR downregulation (Figure 7D). Collectively, these results demonstrate that in diet-induced NASH mice, pharmacological activation of FXR by tropifexor reduced fibrosis, apoptosis, and inflammation, which was dependent, at least in part, on the induction of FincoR.

FincoR expression is increased in chronic liver disease with hepatic inflammation and liver injury

To determine whether expression of FincoR is altered in chronic liver disease, we utilized mouse models of NAFLD/NASH and cholestatic liver injury. Hepatic FincoR levels were significantly increased in mice fed with a high fat diet (HFD) for 12 weeks (Supplementary Figure S1A) and in mice fed a HFD with high fructose (HFHF) in drinking water for 12 weeks (Supplementary Figure S1B). Elevated hepatic FincoR levels were also observed in mice treated with α-naphthylisothiocyanate (ANIT), a chemical inducer of liver cholestasis (Jung et al., 2020; Kim et al., 2016) (Supplementary Figure S1C), and in mice with bile duct ligation (BDL), a surgical method to induce cholestatic liver injury (Supplementary Figure S1D).

The sequence of FincoR is moderately conserved between mice and humans as displayed in the UCSC genome browser (Supplementary Figure S1E). Annotation in the NCBI genome data viewer of the human sequence region with similarity to mouse FincoR revealed an functionally uncharacterized human lncRNA, XR_007061585.1, in this region (Supplementary Figure S1F). However, whether this sequence (or an as yet to be identified lncRNA) is functional or not has not been determined. To explore the potential changes or role of lncRNA XR_007061585.1 in human liver pathological conditions, we measured hepatic levels of the transcripts in PBC and NAFLD patients. Compared to normal individuals, hepatic lncRNA XR_007061585.1 levels were elevated in patients with PBC or NAFLD, but not in severe NASH-fibrosis patients (Supplementary Figure S1G, H). These results demonstrate that hepatic levels of a potential human analog of FincoR are elevated in NAFLD and PBC patients, as FincoR is in mouse models of chronic liver disease with hepatic inflammation and liver injury. However, whether human lncRNA XR_007061585.1 is analogous to mouse FincoR in terms of functions and mechanisms, and whether elevated XR_007061585.1 levels may have a role in the disease progression or may be an adaptive response to liver injury remains to be determined.

Discussion

FXR maintains metabolic homeostasis by transcriptional regulation of genes. Direct regulation of protein-coding genes by FXR, including Shp, is well characterized. In this study we show that FXR also mediates its functions by induction of lncRNA genes, which vastly outnumber protein-coding genes. We further show that pharmacological activation of FXR by hammerhead-type agonists induces a liver-specific enhancer-derived lncRNA, which we named FincoR, that contributes to reduction of NASH pathologies in mice.

FincoR is specifically induced by the hammerhead class of FXR agonists, such as GW4064, cilofexor, and tropifexor. GW4064 is the mother compound of these isoxazole-type hammerhead ligands but is not an ideal therapeutic agent because of its poor water solubility and pharmacokinetics (Abel et al., 2010). In contrast, tropifexor and cilofexor have better pharmacokinetics and are generally well-tolerated in clinical trials for NASH and PBC patients (Abenavoli et al., 2018; Kremoser, 2021; Sanyal et al., 2023; Tully et al., 2017), but the underlying mechanisms for their beneficial effects are poorly understood. Intriguingly, a recent study showed that the gene signature regulated by tropifexor-activated FXR appears to be broader than that of OCA, partly because the tropifexor backbone allows a more favorable interaction of FXR with coactivators or epigenomic modulators (Hernandez et al., 2019). Further, tropifexor was shown to regulate distinct sets of genes in experimental NASH as compared to other FXR agonists, particularly genes involved in fibrosis, inflammation, and oxidative stress (Hernandez et al., 2019). Utilizing CRISPR/Cas9-mediated liver-specific knockdown of FincoR in diet-induced NASH mice, we demonstrate that beneficial effects on reducing liver fibrosis, inflammation, and apoptosis mediated by tropifexor were largely dependent on FincoR (Model, Figure 7E).

While this study focused on regulation of FincoR by pharmacological activation of FXR, physiological and pathological regulations of FincoR appear to be complex. Although FincoR can be induced by the hammerhead class of FXR agonists, it was not induced by the endogenous FXR ligand, cholic acid. This implies that FincoR may not contribute to the physiological functions of FXR. In an effort to investigate the potential role of FincoR in the pathological conditions, we observed that FincoR levels were elevated in mouse models of cholestasis and NASH as well as in human PBC and NAFLD patients, where BA metabolism is dysregulated. Since different BAs can activate or repress the gene-regulating function of FXR (Wahlstrom et al., 2016), altered BA composition in these pathological conditions may contribute to induction of FincoR. Further, binding peaks for multiple nuclear receptors, FXR, LXR, PPARα, RXRα, and HNF-4α, were detected in the FincoR locus so that regulation of FincoR likely involves the combinatorial regulation by multiple nuclear receptors (Supplementary Figure S2). Interestingly, occupancy of the nuclear receptor PPARα at the enhancer region was increased in fasted mice (Supplementary Figure S3). It will be interesting to investigate whether and how FincoR is differently regulated by these nuclear receptors in response to physiological and pathological cues.

The roles of regulatory RNAs in liver function and diseases and their potentials as therapeutic targets are increasingly being appreciated (Brocker et al., 2020; Li et al., 2021; Sallam et al., 2016; Zhao et al., 2018). For example, a critical role for an oxysterol nuclear receptor LXR-induced lncRNA, LeXis, in feedback modulation of cholesterol biosynthesis has been shown (Sallam et al., 2016). Recently, the role of a lncRNA, Pair, in liver phenylalanine metabolism has been demonstrated (Li et al., 2021). Enhancer RNAs are a less characterized class of lncRNAs and are highly associated with enhancer functions in gene regulation (Li et al., 2016). Numerous studies have revealed transcriptional roles for eRNAs in various cellular processes (Lai & Shiekhattar, 2014; Lam et al., 2014; Li et al., 2016; Sartorelli & Lauberth, 2020) but most previous eRNA studies have used cultured cells (Hsieh et al., 2014; Li et al., 2013), with only a few in vivo studies in mouse models (Mirtschink et al., 2019; Tang et al., 2023). In our current study, through integrative analysis of transcriptome and histone mark ChIP-Seq, we identified a group of FXR- regulated eRNAs, including the highly induced FincoR. Our current work characterized the role of FincoR in gene regulation and in mediating beneficial pharmacological effects of tropifexor in NASH, representing important progress in understanding the roles of eRNAs in vivo. Future work is warranted to elucidate the exact mechanisms by which FincoR facilitates action of FXR agonists to alleviate inflammation, fibrosis and apoptosis. RNA inside the cells usually associates with different RNA-binding proteins (RBPs) (Gerstberger et al., 2014). We identified potential binding proteins of FincoR using the ATtRACT database (Giudice et al., 2016). The top four candidates for FincoR binding are KHDRBS1, RBM38, YBX2 and YBX3 (Supplemental Table S5). KHDRBS1 and RBM38 have been reported to have important roles in RNA processing (Bielli et al., 2011; Zou et al., 2021). YBX2 and YBX3 belong to the Y-box (YBX) protein family, which have been linked to diverse forms of RNA metabolism and many other processes, including cell proliferation, DNA repair, stress responses, development, and inflammation (Kleene, 2018). Whether these predicted RBPs interact with FincoR and how they contribute to phenotypes should be investigated in future experimentation to understand the mechanisms involved in FincoR-regulated hepatocyte function.

There is limitation of some of our approaches that cannot fully dissect the underlying mechanisms of FincoR. Currently, the function of eRNA loci can be attributed to a functional eRNA transcript, or binding of transcription factors to this region, or the transcription process itself, or some combination of these effects (Li et al., 2016; Sartorelli & Lauberth, 2020). In our studies to decrease expression of FincoR, the region containing the FXR binding site was deleted (Figure 5-figure supplement 1A, right panel). While the expression of FincoR transcripts was significantly reduced, we cannot rule out whether decreased binding of FXR or decreased transcription contribute to the observed changes in phenotype. To directly investigate the function of FincoR transcripts, downregulation of the transcript with antisense oligonucleotides together with overexpression of FincoR in the liver will be required in future experiments.

FXR is increasingly recognized as an important therapeutic target for enterohepatic diseases, but the development of clinically applicable and more targeted FXR-based therapy is still challenging. In this study, we provided the first characterization of an eRNA, FincoR, induced by pharmacological activation of FXR and show that FincoR has a beneficial role in reducing liver fibrosis and inflammation in dietary NASH mice. Complete understanding of the function and mechanisms of FincoR may provide novel insights for the development of desirable therapy for NASH and other chronic liver diseases.

Materials and Methods

Animal experiments

All animal studies were performed according to procedures approved by the Institutional Animal Care & Use Committee at the University of Illinois at Urbana-Champaign and were in accordance with National Institutes of Health guidelines. Mice were maintained in 12/12 h light/dark cycles and fed standard rodent chow. FXR-LKO mice were generated by breeding FXR floxed mice with Albumin-Cre mice (The Jackson Lab). FincoR-LKD mice were generated as previously reported (Zhao et al., 2018). Briefly, Cas9 transgenic mice (JAX #024858) were injected via the tail vein with adenoviruses (approximately 5×108 PFU) expressing two sgRNAs targeting FincoR (sgRNA1: GGGTTAAGAGCTGTAGGCTG and sgRNA2: ACTTCTATGTCCAACAACCG). The sequences of sgRNAs were designed using a CRISPR design tool (http://crispr.mit.edu/).

Mice were given a single dose of vehicle or 30 mg/kg GW4064 (in corn oil, Tocris Bioscience, #2473) after overnight fasting. Mice were treated with 0.5 mg/kg tropifexor (in corn oil, MedChem Express, HY-107418), 30 mg/kg cilofexor (in corn oil, MedChemExpress, HY-109083), 100 mg/kg fexaramine (in 0.5% methylcellulose, MedChem Express, HY-10912), and 20 mg/kg OCA (in 0.5% methylcellulose, MedChem Express, HY-12222) as indicated. C57BL6 mice were fed with a chow diet containing 0.5% CA for 6 h (Jung et al., 2020).

To induce cholestasis, mice were treated by gavage with 75 mg/kg α-naphthyl isothiocyanate (ANIT) for 48 hours as previously reported (Jung et al., 2020; Kim et al., 2016; Kim et al., 2020). Cholestasis was also induced by bile duct ligation or sham operation in mice for 24 hours or 72 hours (Li et al., 2018).

To induce dietary obesity, mice were fed a high-fat diet (TD88137; Harlan Teklad) or a high-fat diet with 25% fructose in water (high fat/high fructose) for 12 weeks (Seok et al., 2021).

To investigate the effect of liver-specific downregulation of FincoR on NASH, male Cas9 mice were fed the AMLN diet (Research Diets, D09100310, 40 kcal% fat, 2% cholesterol, 20 kcal% fructose) for 12 weeks and then, were injected with adenovirus expressing control sgRNA or sgRNA targeting FincoR. Administration of tropifexor was started 3 days later and given at 0.3 mg/kg dissolved in corn oil.

RNA-Seq

C57BL/6 mice were fasted overnight and i.p. injected with vehicle or GW4064 (30 mg/kg) for 1 h, and livers were collected (n = 4 mice for either vehicle or treated group). Total RNA from each liver was extracted by RNeasy kit (Qiagen), and two randomly selected mice liver RNAs were pooled for RNA-seq (two RNA-seq reactions from 4 mice livers for either vehicle or treated group). RNA-seq was performed as previously described (Byun et al., 2018; Byun et al., 2020; Seok et al., 2018). Ribosomal RNA was removed with the Ribozero HMR Gold kit (Illumina). The sequencing library was generated following methods described below.

Construction of strand-specific RNAseq libraries

Construction of the RNAseq libraries and sequencing on the Illumina NovaSeq 6000 were performed at the Roy J. Carver Biotechnology Center at the University of Illinois at Urbana-Champaign. After DNase digestion, purified total RNAs were analyzed on a Fragment Analyzer (Agilent) to evaluate RNA integrity. The total RNAs were converted into individually barcoded polyadenylated mRNAseq libraries with the Kapa HyperPrep mRNA kit (Roche). Libraries were barcoded with Unique Dual Indexes (UDI’s) which have been developed to prevent index switching. The adaptor-ligated double-stranded cDNAs were amplified by PCR for 8 cycles with the Kapa HiFi polymerase (Roche). The final libraries were quantitated with Qubit (Thermo Fisher) and the average cDNA fragment sizes were determined on a Fragment Analyzer. The libraries were diluted to 10 nM and further quantitated by qPCR on a CFX Connect Real-Time qPCR system (Bio-Rad) for accurate pooling of barcoded libraries and maximization of number of clusters in the flowcell.

Sequencing of libraries in the NovaSeq

The barcoded RNAseq libraries were loaded on one SP lane on a NovaSeq 6000 for cluster formation and sequencing. The libraries were sequenced from one end of the fragments for a total of 100 bp. The fastq read files were generated and demultiplexed with the bcl2fastq v2.20 Conversion Software (Illumina, San Diego, CA). The quality of the demultiplexed fastq files was evaluated with the FastQC software, which generates reports with the quality scores, base composition, k-mer, GC and N contents, sequence duplication levels and overrepresented sequences.

GRO-seq

To harvest the nuclei from mouse liver cells, the liver was harvested at indicated time and washed with a cold swelling buffer (10 mM Tris-HCl, pH 7.5, 2 mM MgCl2, 3 mM CaCl2, 2 U/ml Superase-In). The nuclei were prepared by Dounce homogenization in cold swelling buffer and filtered using a cell strainer (100 μm, BD Biosciences). Nuclei were collected by centrifugation at 400 x g for 10 min, then resuspended in the lysis buffer (swelling buffer with 10% glycerol and 1% IGEPAL) and incubated on ice for 5 min. Nuclei were washed twice with the lysis buffer and resuspended at a concentration of 108 nuclei/ml in the freezing buffer (50 mM Tris-HCl, pH 8.3, 40% glycerol, 5 mM MgCl2, 0.1 mM EDTA). We then followed our previous method (Li et al., 2013; Oh et al., 2021) to conduct nuclear run-on and GRO-Seq library preparation. Briefly, the nuclei in freezing buffer were subjected to the nuclear run-on reaction by mixing with an equal volume of run-on buffer (10 mM Tris-HCl, pH 8.0, 5 mM MgCl2, 1 mM dithiothreitol (DTT), 300 mM KCl, 20 units of Superase-In, 1% sarkosyl, 500 µM ATP, GTP, Br-UTP and 2 µM CTP) for 5 min at 30 °C. The BrU-labeled run-on RNAs were extracted by TRIzol and purified by anti-BrdU agarose beads (Santa Cruz Biotech, sc-32323 AC). The run-on RNAs were then subjected to end repair by T4 PNK, poly-adenylation, and then to cDNA first strand synthesis by a custom primer (oNTI223) that allows circularization of the cDNA. This cDNA then was re-linearized by Ape1 (NEB), size selected by TBE gel electrophoresis and the products of the desired size were excised (∼320–350 bp) for final library prep and sequencing. GRO-Seq samples were run on a NextSeq 500 sequencer from Illumina with a single-end 80 nt model.

Histological analyses

For histology, tissues were dissected and immediately fixed in 10% formalin overnight and processed for paraffin embedding and H&E staining. Paraffin-embedded liver sections were incubated with F4/80 antibody, and antibody was detected using a peroxidase-based method (Abcam, ab64238). Liver collagen was detected by Sirius Red staining (Abcam, ab246832) and apoptosis was detected by TUNEL staining (Millipore, S7100). For Oil Red O staining, liver tissue was frozen in OCT compound (Sakura Finetek, 4583), sectioned, and stained. Liver sections were imaged with a NanoZoomer Scanner (Hamamatsu) and quantification was done using NIH ImageJ.

Metabolic analyses

Hepatic levels of TG (Sigma, MAK266), cholesterol (Sigma, MAK043), glycogen (Biovision, K646-100), total BA levels (Diazyme, DZ042A), serum NEFA (Sigma, MAK044), serum ALT (Sigma, MAK052) and serum AST (Sigma, MAK055) were determined according to the manufacturer’s instructions. Mouse liver IL-1β (R&D systems, MLB00C) and CCL2 (R&D systems, DY479-05) were detected by commercially available ELISA kit.

Liver samples of PBC and NAFLD patients

Liver specimens from normal organ donors, and patients with PBC or NAFLD were obtained from the Liver Tissue Procurement and Distribution System. The samples were unidentifiable, and thus, ethical approval was not required. Hepatic XR_007061585.1 levels were measured by qPCR.

Mouse liver ChIP

Liver ChIP assay was performed as described previously (Jung et al., 2020; Seok et al., 2018). Briefly, chromatin extracts were prepared from FXR floxed and FXR-LKO mouse livers after treatment of the mice with GW4064 which was followed by preclearing and immunoprecipitation using control IgG or FXR antibody (Novus Biologicals, NBP2-16550; Santa Cruz, sc-25309), RXRα antibody (Proteintech, catalog no. 21218-1-AP), or BRD4 antibody (Bethyl Laboratories, catalog #A301-985A50). Enrichment in chromatin precipitates of gene sequences was measured by qRCR using primers listed (Supplementary Table S6).

Primary mouse hepatocytes and HepG2 cells

Primary hepatocytes were isolated from C57BL/6 mice by collagenase (Worthington Biochemical Corporation, LS004188) perfusion and maintained in William’s E Medium with primary hepatocyte maintenance supplements (Gibco #CM4000) in six-well plates as described previously (Jung et al., 2020; Kim et al., 2015). HepG2 cells were maintained in DMEM supplemented with 10% fetal bovine serum at 37°C and 5% CO2.

Luciferase reporter assay

The enhancer region containing the FXR binding element was amplified by PCR and cloned into the pGL4.23 vector (Promega). HepG2 cells were transiently transfected with FXR (100 ng/well) and RXRα (5 ng/well) in combination with reporters containing the wild type FXRE or mutated FXRE (200 ng/well) (Supplementary Table S6). β-galactosidase plasmid (200 ng/well) was also transfected as an internal control. Cells were treated with GW4064 or DMSO after transient transfection. Six hours later, the cells were harvested. All reporter assays were repeated at least three times in triplicates.

RACE

5’ and 3’ RACE assays were performed using a SMARTer RACE kit (Clontech) according to the manufacturer’s instructions. The resulting PCR products were separated by electrophoresis in agarose gels and cloned into the pRACE vector provided by the kit. The transcription start sites and end sites of FincoR were determined by sequencing. The gene-specific primers used for 5’- and 3’-RACE are listed in Supplementary Table S6.

In vitro transcription and translation (TNT)

Expression plasmids for luciferase and FincoR were mixed with a Coupled Reticulocyte Lysate System (Promega). After incubating at 30°C for 60 min, translated products were separated on 4-20% gradient SDS polyacrylamide gels and transferred to PVDF membranes. Chemiluminescent detection of in vitro translated protein was performed following the manufacturer’s protocol (Promega).

RT-qPCR

Total RNA was extracted using the RNeasy Mini Kit (Qiagen, 74104) and 2 µg of RNA was reverse transcribed and mRNA expression was normalized relative to that of 36B4. The qPCR primers are shown in Supplementary Table S6.

Subcellular fractionation

Using a Cytoplasmic and Nuclear RNA Purification Kit (Norgen, Thorold, ON, Canada), the cytoplasm and nucleus fractions from primary hepatocytes isolated from livers of mice treated with GW4064 or DMSO were separated, and both fractions were subjected to RNA extraction and qRT-PCR.

Immunoblotting analysis

Total liver lysates were prepared as described before (Sun et al., 2022). Antibodies for immunoblotting for 𝛽-ACTIN (#4970) were purchased from Cell Signaling Technology. Antibodies for immunoblotting for FXR (sc-25309) were purchased from Santa Cruz.

FincoR polyadenylation study

Total RNA was prepared using the RNeasy mini prep kit (Qiagen). Poly(A)+ and poly(A)-RNA was separated using a Dynabeads™ mRNA Purification Kit (ThermoFisher Scientific) according to the manufacturer’s instructions. Briefly, total RNA was incubated with the Dynabeads/binding buffer suspension at room temperature for 5 min and the reaction tubes were placed on a magnet until solution was clear. The supernatant containing poly(A)-RNA was saved. The beads with poly(A)+ RNA were washed three times with washing buffer provided by the kit. RNA was extracted from the supernatant and beads respectively using Trizol. qPCR was performed to analyze levels of FincoR and 36b4 in the poly(A)+ and poly(A)-RNA fractions.

Genomics Analysis

RNA-Seq and GRO-Seq reads were mapped to the mouse reference genome mm10 with STAR aligner (Dobin et al., 2013). Transcript quantifications were done with the HOMER tool set (Heinz et al., 2010) and enhancer RNAs were identified based on H3K27ac ChIP-Seq in mouse liver. Briefly, intergenic H3K27ac ChIP-Seq peaks were selected as putative enhancer regions in mouse liver. Then +/- 3kb regions around putative enhancers with RNA-Seq signal (>1 RPKM) were considered as putative eRNAs in mouse livers. Overlapped eRNAs regions were merged, and redundant ones were removed. In addition, any +/- 3kb extended eRNAs regions that overlapped with protein-coding genes were further removed to avoid transcriptional readthrough from genes. We used DESeq2 (Love et al., 2014) to identify significantly regulated eRNAs with a cutoff of (FDR < 0.05, Log2FC > 1). Public ChIP-Seq datasets were obtained from ENCODE or GEO (Supplementary Table S3).

Statistical analysis

Statistical analysis was performed using GraphPad Prism 9. Statistical differences were evaluated using the two-tailed unpaired Student’s t-test for comparisons between two groups, or ANOVA and appropriate post hoc analyses for comparisons of more than two groups. Statistical methods and corresponding p values for data shown in each panel are indicated in figure legends.

Data availability

RNA-seq and Gro-seq data were deposited in GEO under the following accession numbers GSE 221986.